---
title: "Syllabus Marketing Research and Analytics Module A AE"
author:
  - name: Luis Francisco Gómez López
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Universidad Militar Nueva Granada
date: 2024-01-22
format:
  html: 
    number-sections: true
    toc: true
    embed-resources: true
bibliography: r_for_marketing_research_and_analytics.bib
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo    = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r libraries}
library(tidyverse)
library(gt)
```

```{css}
p {
  text-align: justify;
}
```

# Professor and meeting hours with students

-   **Name**

    -   Luis Francisco Gomez López

-   **Communication through the virtual classroom**

    -   Every Saturday between 08:00 hours and 10:00 hours:

        -   *Primer Corte 30% \> Virtual Meetings \> Meeting hours with students*

    -   Doubts and concerns of the topics of the course, please use the *Forum: Questions or concerns* that can be accessed at the top of the platform

        -   Talk about `reprex` and **minimal reproducible example**: [**Pending**]{style="color: red;"}

            -   Check this: <https://matthewrocklin.com/minimal-bug-reports.html>

        -   [**Pending**]{style="color: red;"}: In this way, doubts can be shared with all students and all course participants can contribute to the learning process

    -   Complaints regarding grades, please send an email to [**luis.gomezl\@unimilitar.edu.co**](mailto:luis.gomezl@unimilitar.edu.co){.email}

        -   To make any complaints, you are asked to wait first for the grades to be published on the News Bulletin Board and to be made based on the feedbacks that will be made taking into account the evaluation parameters

        -   Urgent issues, you can call me to +57 3124541536 Monday, Wednesday and Friday between 16:00 hours and 17:00 hours. Please do not send me whatsapp messages or call me on weekends.

# Presentation and objectives of the subject

## Description

-   Credits: 4
-   Type: Theoretical-Practice
-   Area of Knowledge: Professional
-   Semester: 1
-   Block: 1 (April 1 - June 4, 2024)

## Justification

The ability to analyze data and make decisions based on evidence is crucial in today's business world. Professionals in business administration need to be familiar with data analysis tools and techniques to better understand their customers, make informed decisions, and improve the profitability of their companies.

This course will provide students with an understanding of data analysis and marketing. The course will cover the basics of programming, statistics, and data analysis. Students will also learn how to apply these skills to solve problems in marketing and how to use a programming language to perform visualizations. By the end of the course, students will have the ability to use a programming language as a data analysis tool and apply their skills in a business context.

## General objective

Develop skills in programming, statistics, and data analysis to solve business problems related to marketing.

# Competencies

-   Global competencies:

    -   Apply statistical analysis techniques and programming in the evaluation of market data to identify trends and opportunities for improvement in products and commercial strategies.

-   Specific competencies:

    -   Prepare a reproducible market report by using statistical analysis techniques and programming to identify trends and opportunities for improvement in products and commercial strategies.

-   Competencies SABE curriculum

    -   Ability to use digital technologies effectively and efficiently
    -   Skills in handling a second language
    -   Ability to use the scientific method and logical thinking

# Schedule and contents

## Reading schedule

```{r}
# Include a footnote for the optional sections
# Generate a section for the exercises
# See https://github.com/quarto-dev/quarto-cli/issues/3340
tibble(Week = c(1:9),
       Material       = c(rep("[@chapman_r_2019]", 
                              times = 9)),
       `Chapter`        = c("1", "2", "3", "4", "5", "6", "7", "8", "11")) |> 
  gt() |> 
  fmt_markdown() |> 
  as_raw_html()
```

## Detailed schedule of virtual meetings and activities

```{r}
islands_tbl <- 
  tibble(
    name = names(islands),
    size = islands
  ) |>
  arrange(desc(size)) |>
  slice(1:10)

islands_tbl |> 
  gt()
```

### Primer corte 30%

-   April 1

    -   Start Block 2

    -   **Virtual meeting Presentation Syllabus**

        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Review the course program in advance, which can be downloaded by following the route *Primer corte 30% \> Learning Resources \> Support material \> Presentations \> 000_syllabus_intro_econ_2024_1*
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   January 23

    -   **Virtual meeting Introduction**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 1]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   January 29

    -   **Questionnaire 1 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 1 & 2]

-   January 30

    -   **Virtual meeting Production and Income I**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 1

    -   **Virtual meeting Production and Income II**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 4

    -   **Questionnaire 1 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 1 before 22:00 hours

-   February 6

    -   **Virtual meeting Economic Growth**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 10

    -   **Oral presentation Midterm Exam 1 Opening**
        -   **Opening time**: 16:00 hours
        -   **Closing time**: 18:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Segundo corte 30%

-   February 12

    -   **Questionnaire 2 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 4 & 5]

-   February 13

    -   **Virtual meeting Economy and institutions**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 4]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 20

    -   **Virtual meeting External sector I**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 22

    -   **Virtual meeting External sector II**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   February 25

    -   **Questionnaire 2 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before 22:00 hours

-   February 27

    -   **Virtual meeting Public Finances**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   March 2

    -   **Oral presentation Midterm Exam 2 Opening**
        -   **Opening time**: 16:00 hours
        -   **Closing time**: 18:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Tercer corte 40%

-   March 4

    -   **Questionnaire 3 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 7 & 8]

-   March 5

    -   **Virtual meeting Money, prices and the exchange rate**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 7]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   March 12

    -   **Virtual meeting Financial Market**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 8]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   March 17

    -   **Questionnaire 3 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before 22:00 hours

-   March 19

    -   **Virtual meeting Labor Market**
        -   **Meeting time**: 19:00 - 19:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   June 1

    -   **Oral presentation Final Exam Opening**
        -   **Opening time**: 16:00 hours
        -   **Closing time**: 18:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

# Qualifiable activities, weights and process in case of no presentation

## Activities and weights

-   **Primer corte**: 30%

    -   **Questionnaire 1**: 15%
    -   **Oral presentation Midterm Exam 1**: 15%

-   **Segundo corte**: 30%

    -   **Questionnaire 2**: 15%
    -   **Oral presentation Midterm Exam 2**: 15%

-   **Tercer corte**: 40%

    -   **Questionnaire 3**: 15%
    -   **Oral presentation Final Exam**: 25 %

## Process in case of no presentation

-   Students who do not present a **Questionare** due to force majeure or illness, must send the respective excuse scanned through the [**luis.gomezl\@unimilitar.edu.co**](mailto:luis.gomezl@unimilitar.edu.co){.email}. If it is not sent, the note will be zero. Once the excuse has been satisfactorily received, the student should **not** present the activity and the corresponding grade will be equal to the immediately subsequent grade obtained in the **Midterm Exam 1** or **Midterm Exam 2** or **Final Exam** according to the case.

-   Students who do not present a **Midterm Exam 1** or **Midterm Exam 2** or the **Final Exam**, must send and email to [**admon.distancia\@unimilitar.edu.co**](mailto:admon.distancia@unimilitar.edu.co){.email} two days before or two days after the closing date of these activities. However, this request can be made only once during the course. If it is not requested under these conditions, the grade will be zero.

# Use of Chat GPT-3, PaLM 2 (used in BART) and its friends

You can use any artificial intelligence tool and in particular any Large Language Model but please use them in a responsible way by taking into account its capabilities and limitations and always pointing out to the professor and your colleges that you have use them

# Study material

-   [@chapman_r_2019, Chapters 1-8, 11]

    -   Complete route:

        -   <https://www.umng.edu.co/inicio> \> Biblioteca \> Bases virtuales y libros electrónicos \> S \> Springer Link \> Ingresar usuario y contraseña [^1] \> Search "R For Marketing Research and Analytics" \> R For Marketing Research and Analytics Book 2019 [^2]

    -   Direct link:

        -   <https://doi-org.ezproxy.umng.edu.co/10.1007/978-3-030-14316-9>

[^1]: Remember that the user and password to access and read the e-book are the same that you use to check your grades

[^2]: Download the book in your preferred format from the available options

```{r}
# include the part about the site of the book
# https://r-marketing.r-forge.r-project.org/index.html
```

# About plagiarism

-   **Acuerdo 02 de 2015 (7 de Mayo): Artículo 68. Escala de Calificaciones Numeral f**

*"La comprobación de todo fraude o intento de fraude en cualquier tipo de prueba, acarreará al estudiante y a su(s) colaborador(es), una calificación de CERO PUNTO CERO (0.0) en la prueba respectiva. Como quiera que el fraude o intento de cometerlo es una falta grave, el docente debe informar de este hecho al Director del programa respectivo, quien enviará el informe a la Decanatura de la Facultad, con el fin de iniciar los trámites correspondientes para aplicar la sanción disciplinaria prevista en éste reglamento.”*

-   **Acuerdo 05 de 2018 (6 de Junio): Artículo 20 que modifica el Artículo 107 Faltas Disciplinarias Numeral c. Número 6 del Acuerdo 02 de 2015 (7 de Mayo)**

*"Hacer plagio en cualquier tipo de documento o desplegar alguna conducta dolosa que viole el régimen de propiedad intelectual de la Universidad."*

# Study material references
