---
title: An overview of the R Language
author: Luis Francisco Gomez Lopez
date: 2023-06-08 23:15:07
format: 
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    number-sections: true
    number-depth: 2
    code-fold: show
    embed-resources: true
editor: visual
bibliography: r_for_marketing_research_and_analytics.bib
---

# Sales and product satisfaction survey

```{r libraries, echo=TRUE, message=FALSE}
library(tidyverse)
library(skimr)
```

-   Ordinal 7 point scale

    -   Extremely satisfied: 7
    -   Moderately satisfied: 6
    -   Slightly satisfied: 5
    -   Neither satisfied or dissatisfied: 4
    -   Slightly dissatisfied: 3
    -   Moderately dissatisfied: 2
    -   Extremely dissatisfied: 1

-   Variables

    -   iProdSAT: satisfaction with a product
    -   iSalesSAT: satisfaction with sales experience
    -   iProdREC: likelihood to recommend the product
    -   iSalesREC: likelihood to recommend the sales person
    -   Segment: market segment assigned by a clustering algorithm

## Import data: the base R way

```{r }
satisfaction_data <- read.csv(file = "http://goo.gl/UDv12g")
satisfaction_data |> head(n=5)
```

## Import data: the tidyverse way

```{r message=FALSE}
satisfaction_data <- read_csv(file = "http://goo.gl/UDv12g")
satisfaction_data |> slice_head(n=5)
```

## Transform data: the base R way

```{r}
satisfaction_data$Segment <- factor(x = satisfaction_data$Segment, 
                                    ordered = FALSE)
summary(object = satisfaction_data)
```

## Transform data: the skimr way (relies on tidyverse)

```{r}
satisfaction_data |> skim()
```
