---
title: "Comparing Groups: Statistical Tests"
format: 
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 4
    include-in-header:
      - text: |
          \usepackage{amsmath}
editor: visual
bibliography: r_for_marketing_research_and_analytics.bib
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(tidymodels)
library(tidyquant)
library(latex2exp)
```

# Preliminaries

## Gamma function

$\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ with $x > 0$[^1] and $t > 0$

[^1]: $\Gamma(x)$ can be defined for $x < 0$ with $x \notin \mathbb{Z}^-$ but we are not interested in those cases

```{r}
#| echo: false
#| fig-align: center
#| warning: false
ggplot() +
  geom_function(fun = gamma, 
                xlim = c(0,5),
                color='#2C3E50') +
  labs(x=NULL,
       y=NULL,
       title = 'Gamma function') +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

### Properties Gamma function

$$\begin{split}
 \Gamma(\frac{1}{2}) & = \int_0^\infty t^{-\frac{1}{2}}e^{-t}dt \\
 & = \int_0^\infty \frac{\sqrt{2}}{z}e^{-\frac{z^2}{2}}zdz \text{ with } t = \frac{z^2}{2} \\
 & = \int_0^\infty \sqrt{2}\sqrt{2}\sqrt{\pi}\frac{1}{\sqrt{2\pi}}e^{-\frac{z^2}{2}}dz \\
 & = 2\sqrt{\pi} \int_0^\infty \frac{1}{\sqrt{2\pi}}e^{-\frac{z^2}{2}}dz \\
 & = 2\sqrt{\pi} \frac{1}{2} \\
 & = \sqrt{\pi} \\
 \end{split}$$

$$\begin{split}
  \Gamma(x+1) & = \int_0^\infty t^xe^{-t}dt \\
  & = [-t^xe^{-t}]_{t=0}^{t=\infty} + x\int_0^\infty t^{x-1}e^{-t}dt \text{ where we apply integration by parts} \\
  & = -\lim_{x \to \infty} \frac{t^x}{e^t} + x\Gamma(x) \\
  & = x\Gamma(x) \text{ where we apply L'Hôpital's rule several times}
  \end{split}$$

In the case of $x \in \mathbb{N}^*$ we can show that $\Gamma(x+1) = x!$

-   For $x=2$ we have that $\Gamma(2) = 2\Gamma(1) = 2\int_0^\infty e^{-t} = 2[-e^{-x}]_{x=0}^{x=\infty} = 2\cdot1$
-   Assume $\Gamma(x) = (x-1)!$
-   Let $\Gamma(x+1) = x\Gamma(x) = x(x-1)! = x!$

## Gamma distribution function

$f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}$ with $x > 0$, $\alpha > 0$ and $\beta > 0$

```{r}
#| echo: false
#| fig-align: center
ggplot() + 
  geom_function(fun=dgamma, args=list(shape=0.5, rate=0.5),
                xlim=c(0,15),
                color='#2C3E50') + 
  scale_x_continuous(breaks = 0:15) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Gamma distribution function',
       subtitle = TeX(r'($\alpha=\frac{1}{2}, \beta=\frac{1}{2}$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

Let $f(x;\frac{1}{2},\frac{1}{2}) = \frac{\frac{1}{2}^{\frac{1}{2}}}{\Gamma(\frac{1}{2})}x^{\frac{1}{2}-1}e^{-\frac{x}{2}} = \frac{1}{\sqrt{2\pi}}x^{-\frac{1}{2}}e^{-\frac{x}{2}}$. Suppose that $Z \sim \mathcal{N}(0,1)$ then it is possible to show that $Z^2 \sim f(x;\frac{1}{2},\frac{1}{2})$

If $F_{X^2}$ is the cumulative distribution function of $X^2$ then for any $x \leq 0$ we have that $F_{X^2} = \mathbf{P}(X^2 \leq x) = 0$.

In the case of $x > 0$ we have that $F_{X^2} = \mathbf{P}(X^2 \leq x) = \mathbf{P}(-\sqrt{x} \leq X \leq \sqrt{x}) = 2\mathbf{P}(X < \sqrt{x}) = 2F_X(\sqrt{x})$

We can recover $f_{X^2}$ taking into account that $f_{X^2}(x) = \frac{dF_{X^2}(x)}{dx}$. For $x \leq 0$ we have that $f_{X^2} = 0$ and for $x > 0$ we have that $\frac{dF_{X^2}(x)}{dx} = \frac{d2F_X(\sqrt{x})}{dx} = 2\frac{dF_X(\sqrt{x})}{dx} = 2f_X(\sqrt{x})\frac{1}{2\sqrt{x}} = f_X(\sqrt{x})x^{-\frac{1}{2}} = \frac{1}{\sqrt{2\pi}}e^{-\frac{x}{2}}x^{-\frac{1}{2}} = f_{X^2}(x)$

### Definition of a chi-squared distribution function

A random variable $X$ follows a chi-squared distribution with $k$ degrees of freedom if it its distribution function is:

$$f(x; \frac{k}{2}, \frac{1}{2}) = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})}x^{\frac{k}{2}-1}e^{-\frac{x}{2}}$$ Where $k > 0$ and $k \in \mathbb{N}$. We will use the notation $X \sim \chi^2(k)$

### Moment generating function

The moment generating function of a random variable $X$ is $M_{X}(t) = E[e^{tX}] = \int_{-\infty}^{\infty} e^{tx} f(x) dx$

In the case of $X \sim \chi^2(k)$ we have that:

$$\begin{split}
  M_X(t) & = \int_{0}^{\infty} e^{tx} \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})}x^{\frac{k}{2}-1}e^{-\frac{x}{2}} dx \\
  & = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})} \int_{0}^{\infty} x^{\frac{k}{2}-1} e^{-x(\frac{1}{2} - t)} dx \\ 
  \end{split}$$

If $\frac{1}{2} - t < 0$ then $e^{-x(\frac{1}{2} - t)} \longrightarrow \infty$ so the integral diverges in this case and the expectation fails to exist.

If $\frac{1}{2} - t = 0$ then $\int_{0}^{\infty} x^{\frac{k}{2}-1} dx$ but this means that $\int_{0}^{1} x^{\frac{k}{2}-1} dx + \int_{1}^{\infty} x^{\frac{k}{2}-1} dx$ don't exist because $\int_{1}^{\infty} x^{\frac{k}{2}-1} dx$ don't exist taking into account that $\frac{k}{2} - 1 \geq -\frac{1}{2}$

If $\frac{1}{2} - t > 0$ then $\frac{1}{2} > t$. We can define $u=\frac{1}{2} - t$ so

$$\begin{split}
  M_X(t) & = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})} \int_{0}^{\infty} x^{\frac{k}{2}-1} e^{-x(\frac{1}{2} - t)} dx \\
  & = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})} \int_{0}^{\infty} \biggr( \frac{u}{\frac{1}{2}-t} \biggr)^{\frac{k}{2}-1} \frac{1}{\frac{1}{2} - t} e^{-u} du \text{ with } u=x(\frac{1}{2}-t) \\
  & = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})(\frac{1}{2} - t)^{\frac{k}{2}}} \int_{0}^{\infty} u^{\frac{k}{2}-1} e^{-u} du \\
  & = \frac{1}{2^{\frac{k}{2}}\Gamma(\frac{k}{2})(\frac{1}{2} - t)^{\frac{k}{2}}} \Gamma(\frac{k}{2}) \\
  & = \frac{1}{2^{\frac{k}{2}}\frac{(1 - 2t)^{\frac{k}{2}}}{2^{\frac{k}{2}}}} \\
  & = \frac{1}{(1 - 2t)^{\frac{k}{2}}} \\
  \end{split}$$

## Expected value of a continuous random variable

### Joint density functions

Two random variables $X$ and $Y$ are jointly continuous if there is a function $f_{X,Y}(x,y)$ on $\mathbb{R}^2$ such that

$$\mathbf{P}(X \leq s, Y \leq t) = \int_{-\infty}^t \int_{-\infty}^s f_{X,Y}(x,y) dxdy$$ Such that $f_{X,Y}(x,y) \geq 0$ and $\int_{-\infty}^\infty \int_{-\infty}^\infty f_{X,Y}(x,y) dxdy = 1$

### Marginal densities

If $X$ and $Y$ are jointly continuous with joint density $f_{X,Y}(x,y)$, then the marginal densities are given by

$$f_X(x) = \int_{-\infty}^\infty f_{X,Y}(x,y) dy$$ $$f_Y(y) = \int_{-\infty}^\infty f_{X,Y}(x,y) dx$$

### Definition of independence of random continuous variables

Let $X,Y$ be jointly continuous random variables with joint density $f_{X,Y}(x, y)$ and marginal densities $f_X(x)$, $f_Y (y)$. We say they are independent if

$$f_{X,Y}(x, y) = f_X(x)f_Y (y)$$

### Definition of expected values

If $X$ is a continuous random variable with a probability density function $f(x)$ then

$$E[X] = \int_{-\infty}^{\infty} xf(x)dx$$ If $Z = XY$ where $X$ and $Y$ are continuous independent random variables we have the following result[^2]:

[^2]: This proof may have problems and it is not totally rigurous. For example check out the discrete case in <https://math.stackexchange.com/questions/3091892/proof-of-exy-ex-ey> and the continuous case in <https://math.stackexchange.com/questions/3707726/proof-of-expected-value-property-for-product-of-independent-variables>

$$\begin{split}
  E[Z] & = \int_{-\infty}^{\infty} zf_Z(z)dz \\
  & = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xyf_{X,Y}(x,y) dxdy \\
  & = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xyf_X(x)f_Y(y) dxdy \text{ By independence} \\
  & = \int_{-\infty}^{\infty} xf_X(x)dx \int_{-\infty}^{\infty} yf_Y(y) dy \text{ By Fubini-Tonelli theorem} \\
  & = E[X]E[Y]
  \end{split}$$

If $Y=g(X)$ where $X$ is a continuous random variable with probability density function $f_X(x)$ there is a theorem which specifies that [@rice_mathematical_2021, Chapter 24, page 122]

$$\begin{split}
  E[Y] & = \int_{-\infty}^\infty yf_Y(y)dy \\
  & = \int_{-\infty}^\infty g(x)f_X(x)dx
  \end{split}$$

What happen with $W=g(X)h(Y)$ where $X$ and $Y$ are continuous independent random variables with probability density functions $f_X(x)$ and $f_Y(y)$?

We need to establish if $g(X)$ and $h(Y)$ are also independent random variables. If this is the case we have that $E[W] = E[g(X)]E[h(Y)]$.

Also $E[g(X)]E[h(Y)] = \int_{-\infty}^\infty g(x)f_X(x)dx \int_{-\infty}^\infty h(y)f_Y(y)dy$

## Beta function

$\beta(a,b) = \int_0^1 t^{a-1}(1-t)^{b-1}dt$ where $a > 0$ and $b > 0$

```{r}
#| echo: false
#| fig-align: center
#| warning: false
a <- 1:5
b <- 1:5
z = outer(a, b, beta)
persp(a, b, z,
      theta = 45, phi = 15,
      col = '#2C3E50',
      main = 'Beta function',
      ticktype = 'detailed')
```

We have that:

$$\begin{split}
  \beta(a,b) & = \int_0^1 t^{a-1}(1-t)^{b-1}dt \\
  & = \int_1^0 (1-u)^{a-1}u^{b-1}-du \text{ with } u = 1-t \\
  & = -\int_1^0 (1-u)^{a-1}u^{b-1}du \\
  & = \int_0^1 (1-u)^{a-1}u^{b-1}du \text{ because } \int_b^a f(x) = - \int_a^b f(x) \\
  & = \beta(b,a)
  \end{split}$$

Also because $\Gamma(a) = \int_0^\infty t^{a-1}e^{-t}dt$, $\Gamma(b) = \int_0^\infty t^{b-1}e^{-t}dt$ and $\Gamma(a + b) = \int_0^\infty t^{a + b - 1}e^{-t}dt$ we have that:

$$\begin{split}
  \Gamma(a+b)\beta(a,b) & = \Gamma(a+b) \int_0^1 t^{a-1}(1-t)^{b-1}dt \\
  & = \Gamma(a+b) \int_0^\infty \frac{u^{a-1}}{(1+u)^{a-1}}\frac{1}{(1-u)^{b-1}}\frac{1}{(1+u)^2}du \text{ where } t = \frac{u}{1+u} \\
  & = \Gamma(a+b) \int_0^\infty \frac{u^{a-1}}{(1+u)^{a+b}}du \\
  & = \int_0^\infty v^{a+b-1}e^{-v}dv \int_0^\infty \frac{u^{a-1}}{(1+u)^{a+b}}du \\
  & = \int_0^\infty u^{a-1} \biggl( \int_0^\infty \biggr[ \frac{v}{1+u} \biggr]^{a+b-1} \frac{1}{1+u} e^{-v} dv \biggl) du \\
  & = \int_0^\infty u^{a-1} \biggl( \int_0^\infty s^{a+b-1} e^{-s(1+u)} ds \biggl) du \text{ where } s = \frac{v}{1+u} \\
  & = \int_0^\infty s^b e^{-s} \biggl( \int_0^\infty (us)^{a-1} e^{-us} du \biggl) ds \\
  & = \int_0^\infty s^{b-1} e^{-s} \biggl( \int_0^\infty (us)^{a-1} e^{-us} dus \biggl) ds \text{ where } dus = sdu \\
  & = \Gamma(a) \int_0^\infty s^{b-1} e^{-s} ds \\
  & = \Gamma(a) \Gamma(b) \\
  \beta(a,b) & = \frac{\Gamma(a) \Gamma(b)}{\Gamma(a+b)}
  \end{split}$$

## Incomplete beta function

$\beta(x;a,b) = \int_0^x t^{a-1}(1-t)^{b-1}dt$ where $x \in [0,1]$, $a > 0$ and $b > 0$

## Regularized incomplete beta function

$I_x(a,b) = \frac{\beta(x;a,b)}{\beta(a,b)} = \frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}\int_0^x t^{a-1}(1-t)^{b-1}dt$ where $x \in [0,1]$, $a > 0$ and $b > 0$

$$\begin{split}
  1- I_{1-x}(b,a) & = \frac{\beta(1-x;b,a)}{\beta(b,a)} \\  
  & = 1 - \frac{1}{\beta(b,a)} \int_0^{1-x} t^{b-1}(1-t)^{a-1} dt \\
  & = 1 - \frac{1}{\beta(b,a)} \int_1^x (1-u)^{b-1}u^{a-1} (-du) \text{ where } u = 1-t \\
  & = \frac{\beta(a,b)}{\beta(a,b)} + \frac{1}{\beta(a,b)} \int_1^x u^{a-1}(1-u)^{b-1} dt \\
  & = \frac{1}{\beta(a,b)} \int_0^1 u^{a-1}(1-u)^{b-1} dt + \frac{1}{\beta(a,b)} \int_1^x u^{a-1}(1-u)^{b-1} dt \\
  & = \frac{1}{\beta(a,b)} \int_0^x u^{a-1}(1-u)^{b-1} dt \\
  & = \frac{\beta(x;a,b)}{\beta(a,b)} \\
  & = I_x(a,b) \\
  \end{split}$$

If $p \in [0,1]$, $n \geq x \geq 0$, $n \in \mathbb{N}^*$ and $x \in \mathbb{N}^*$ we can define:

$$\begin{split}
  I_p(x,n-x+1) & = \frac{\beta(p;x,n-x+1)}{\beta(x,n-x+1)} \\
  & = \frac{\Gamma(x+n-x+1)}{\Gamma(x)\Gamma(n-x+1)} \int_0^p t^{x-1}(1-t)^{n-x+1-1} dt \\
  & = \frac{n!}{(x-1)!(n-x)!} \int_0^p t^{x-1}(1-t)^{n-x} dt \\
  & = x\frac{n!}{x!(n-x)!} \int_0^p t^{x-1}(1-t)^{n-x} dt \\
  & = x\binom{n}{x} \int_0^p t^{x-1}(1-t)^{n-x} dt \\
  \frac{dI_p(x,n-x+1)}{dp} & = x\binom{n}{x} \frac{d}{dp} \int_0^p t^{x-1}(1-t)^{n-x} dt \\
  & = x\binom{n}{x} p^{x-1}(1-p)^{n-x} \text{ By the fundamental theorem of calculus} \\
  \end{split}$$

$$\begin{split}
  I_{1-p}(n-x,x+1) & = 1 - I_p(x+1,n-x) \\
  & = 1 - \frac{\beta(p;x+1,n-x)}{\beta(x+1,n-x)} \\
  & = 1 - \frac{\Gamma(x+1+n-x)}{\Gamma(x+1)\Gamma(n-x)} \int_0^p t^x(1-t)^{n-x-1} dt \\
  & = 1 - \frac{n!}{x!(n-x-1)!} \int_0^p t^x(1-t)^{n-x-1} dt \\
  & = 1 - (n-x)\frac{n!}{x!(n-x)!} \int_0^p t^x(1-t)^{n-x-1} dt \\
  & = 1 - (n-x)\binom{n}{x} \int_0^p t^x(1-t)^{n-x-1} dt \\
  \frac{dI_{1-p}(n-x,x+1)}{dp} & = - (n-x)\binom{n}{x} \frac{d}{dp} \int_0^p t^x(1-t)^{n-x-1} dt \text{ By the fundamental theorem of calculus} \\
  & = - (n-x)\binom{n}{x} p^x(1-p)^{n-x-1} \\
  \end{split}$$

## Beta distribution

$f(x;a,b) = \frac{1}{\beta(a,b)}x^{a-1}(1-x)^{b-1} = \frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}x^{a-1}(1-x)^{b-1}$ where $x \in [0,1]$, $a>0$ and $b>0$

```{r}
#| echo: false
#| fig-align: center
ggplot() + 
  geom_function(fun=dbeta, args=list(shape1=2, shape2=5),
                xlim=c(0,1),
                color='#2C3E50') + 
  scale_x_continuous(breaks = seq.int(from = 0, to = 1, by = 0.1)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Beta distribution function',
       subtitle = TeX(r'($a=2, b=5$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

We use the notation $X \sim Beta(a,b)$ to say that the random variable $X$ follows a beta distribution with parameters $a>0$ and $b>0$

### Cumulative distribution function

$$\begin{split}
  F(x; a, b) & = \mathbf{P}(X \leq x) \\ 
  & = \int_{-\infty}^x \frac{1}{\beta(a,b)}t^{a-1}(1-t)^{b-1} dt \\
  & = \frac{1}{\beta(a,b)} \biggr[ \int_{-\infty}^0 t^{a-1}(1-t)^{b-1} dt + \int_{0}^x t^{a-1}(1-t)^{b-1} dt \biggr] \\
    & = \frac{1}{\beta(a,b)} \int_{0}^x t^{a-1}(1-t)^{b-1} dt \\
    & = \frac{\beta(x,a,b)}{\beta(a,b)} \\
    & = I_x(a,b)
  \end{split}$$

## Binomial distribution

A random variable $X$ follows a binomial distribution if its distribution function is:

$f(x; n,p) = \binom{n}{x} p^x(1-p)^{n-x}$

Where $0 < p < 1$, $n \in \{0, 1, \ldots \}$, $x \leq n$ and $\binom{n}{x} = \frac{n!}{x!(n-x)!}$

We use the notation $X \sim B(n,p)$

```{r}
#| echo: false
#| fig-align: center
tibble(x = 0:1,
       y = dbinom(x = x, size=1, prob=0.5)) |> 
  ggplot() + 
  geom_point(aes(x=x, y=y),
             color='black',
             fill='#2C3E50',
             shape=21) + 
  scale_x_continuous(breaks = 0:1) +
  scale_y_continuous(breaks = seq.int(from = 0, to = 1, by = 0.25)) +
  coord_cartesian(ylim = c(0,1)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Bernoulli distribution function',
       subtitle = TeX(r'($p = \frac{1}{2}$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

If $n=1$ we say that $X$ follows a bernoulli distribution:

$f(x; p) = p^x(1-p)^{1-x}$

Where $x \in {0,1}$

We use the notation $X \sim Bernoulli(p)$

### Binomial coefficients

$i \binom{n}{i} = i\frac{n!}{i!(n-i)!} = i\frac{n(n-1)!}{i(i-1)!(n-1-(i-1))!}=n\frac{(n-1)!}{(i-1)!((n-1)-(i-1))!}=n\binom{n-1}{i-1}$

$(n-i)\binom{n}{i} = (n-i)\frac{n!}{i!(n-i)!} = (n-i)\frac{n(n-1)!}{i!(n-i)(n-i-1)!}=n\frac{(n-1)!}{i!((n-1)-i)!}=n\binom{n-1}{i}$

### Cumulative distribution function

$$\begin{split}
  F(x;n,p) & = \mathbf{P}(X \leq x) \\
  & = \sum_{i=0}^x \binom{n}{i} p^i(1-p)^{n-i} \\
  & = \binom{n}{0} p^0(1-p)^{n-0} + \binom{n}{1} p^1(1-p)^{n-1} + \cdots + \binom{n}{x-1} p^{x-1}(1-p)^{n-(x-1)} + \binom{n}{x} p^x(1-p)^{n-x} \\
  & = \binom{n}{0} p^0(1-p)^{n-0} + \binom{n}{1} p^1(1-p)^{n-1} + \cdots + \binom{n}{x-1} p^{x-1}(1-p)^{n-(x-1)} + \binom{n}{x} p^x(1-p)^{n-x} \\
  \frac{dF(x;n,p)}{dp} & = \binom{n}{0} (n-0)p^0(1-p)^{n-1} + \binom{n}{1} (1p^{1-1}(1-p)^{n-1} - (n-1)p^1(1-p)^{n-2}) + \cdots + \\ 
  & \;\;\;\; \binom{n}{x-1} ((x-1)p^{x-2}(1-p)^{n-(x-1)} - (n-(x-1))p^{x-1}p^{n-(x-1)}) + \\
  & \;\;\;\; \binom{n}{x} (xp^{x-1}(1-p)^{n-x} - (n-x)p^x(1-p)^{n-x-1}) \\
  & = \sum_{i=1}^x \binom{n}{i} ip^{i-1}(1-p)^{n-i} - \sum_{i=0}^x \binom{n}{i}(n-i)p^i(1-p)^{n-i-1} \\
  & = n \Biggr[ \sum_{i=1}^x \binom{n-1}{i-1}p^{i-1}(1-p)^{n-i} - \sum_{i=0}^x \binom{n-1}{i}p^i(1-p)^{n-i-1} \Biggr] \\
  & = n \Biggr[ \sum_{i=1}^x \binom{n-1}{i-1}p^{i-1}(1-p)^{n-i} - \sum_{i=1}^x \binom{n-1}{i-1}p^{i-1}(1-p)^{n-i} - \binom{n-1}{x}p^x(1-p)^{n-x-1} \Biggr] \\  
  & = - n\binom{n-1}{x}p^x(1-p)^{n-x-1} \\
  & = - n\frac{(n-1)!}{x!(n-1-x)!}p^x(1-p)^{n-x-1} \\
  & = - (n-x)\binom{n}{x}p^x(1-p)^{n-x-1} \\
  \end{split}$$

Taking into account that $\frac{dF(x;n,p)}{dp} = \frac{dI_{1-p}(n-x,x+1)}{dp}$

#### Relation with the Beta cumulative distribution function

-   According to [@johnson_univariate_2005, page 119] we have the following result:

$$\begin{split}
  Pr[X \geq x] & = \sum_{i=x}^n \binom{n}{i} p^i(1-p_i)^{n-i} \\
  & = I_p(x, n - x + 1) \\
  & = \frac{\beta(p; x, n - x + 1)}{\beta(x, n - x + 1)} \\
  & = \frac{\int_0^p t^{x-1}(1-t)^{n-x}dt}{\int_0^1 t^{x-1}(1-t)^{n-x}dt}
  \end{split}$$

-   Using this result we have also the following result:

$$\begin{split}
  Pr[X \leq x] & = 1 - Pr[X \geq x+1] \\
  & = 1- I_p(x + 1, n - (x + 1) + 1) \\
  & = 1- I_p(x + 1, n - x) \\
  & = 1 - \frac{\beta(p; x + 1, n - x)}{\beta(x + 1, n - x)} \\
  & = 1 - \frac{\int_0^p t^x(1-t)^{n-x-1}dt}{\int_0^1 t^x(1-t)^{n-x-1}dt}
  \end{split}$$

Let's see the application of this result for $Pr[X \geq x]$:

```{r}
#| echo: true
pbinom(q = 156, size = 300, prob = 0.5, 
       lower.tail = FALSE)
pbeta(q = 0.5, shape1 = 157, shape2 = 300 - 157 + 1, 
      lower.tail = TRUE)
```

The first part in relation to $q = 156$ it is fixed in that way taking into account that `lower.tail = FALSE` means that $Pr[X > x]$. Therefore $Pr[X > 156] = Pr[X \geq 157]$

Let's see the application of this result for $Pr[X \leq x]$:

```{r}
#| echo: true
pbinom(q = 157, size = 300, prob = 0.5, 
       lower.tail = TRUE)
1 - pbeta(q = 0.5, shape1 = 157 + 1, shape2 = 300 - 157, 
          lower.tail = TRUE)
```

### Moment generating function

$$\begin{split}
  M_X(t) & = E[e^{tX}] \\
  & = \sum_{x=0}^n e^{tx} \binom{n}{x} p^x (1-p)^{(n-x)} \\
  & = \sum_{x=0}^n \binom{n}{x} (pe^t)^x (1-p)^{(n-x)} \\
  & = (pe^t + 1 - p)^n \text{ By the binomial theorem}
  \end{split}$$

If $X_1, \ldots X_n$ are independent random variables with $X_i \sim Bernoulli(p)$ then:

$$\sum_{i=1}^n X_i \sim B(n,p)$$ We can proof this with the moment generating function

$$\begin{split}
  M_{\sum_{i=1}^n X_i}(t) & = E[e^{t(\sum_{i=1}^n X_i)}] \\
  & = E[e^{tX_1} \ldots e^{tX_n)}] \\
  & = E[e^{tX_1}] \cdots E[e^{tX_n}] \text{ Because } X_1, \ldots X_k \text{ are independent random variables} \\
  & = (pe^t + 1 - p) \ldots (pe^t + 1 - p) \text{ Because } X_i \sim Bernoulli(p) \\
  & = (pe^t + 1 - p)^n
  \end{split}$$

## Normal distribution

A random variable $X$ follows a normal distribution if its distribution function is:

$f(x) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}$

We say that $X \sim \mathcal{N}(\mu, \sigma)$

```{r}
#| echo: false
#| fig-align: center
ggplot() + 
  geom_function(fun=dnorm, args=list(mean=1, sd=0.5),
                xlim=c(-4,6),
                color='#2C3E50') + 
  scale_x_continuous(breaks = seq.int(from = -4, to = 6, by = 1)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Normal distribution function',
       subtitle = TeX(r'($\mu=1, \sigma=\frac{1}{2}$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

### Cumulative distribution function

$$\begin{split}
  F_X(x) & = \int_{-\infty}^x \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2} dx \\
  & = \frac{1}{\sigma\sqrt{2\pi}} \int_{-\infty}^x e^{-(\frac{x-\mu}{\sqrt{2}\sigma})^2} dx \\
  & = \frac{1}{\sigma\sqrt{2\pi}} \int_{-\infty}^{\frac{x-\mu}{\sqrt{2}\sigma}} e^{-t^2} \sqrt{2}\sigma dt \text{ where } t = \frac{x-\mu}{\sqrt{2}\sigma} \\
  & = \frac{1}{\sqrt{\pi}} \int_{-\infty}^{\frac{x-\mu}{\sqrt{2}\sigma}} e^{-t^2} dt \\
  & = \frac{1}{\sqrt{\pi}} \int_{-\infty}^{0} e^{-t^2} dt + \frac{1}{\sqrt{\pi}} \int_{0}^{\frac{x-\mu}{\sqrt{2}\sigma}} e^{-t^2} dt \\
  & = \frac{1}{\sqrt{\pi}} \int_{0}^{\infty} e^{-u^2} du + \frac{1}{\sqrt{\pi}} \int_{0}^{\frac{x-\mu}{\sqrt{2}\sigma}} e^{-t^2} dt \text{ where } t = -u \\
  & = \frac{1}{\sqrt{\pi}} \int_{0}^{\infty} e^{-u^2} du + \frac{1}{\sqrt{\pi}} \int_{0}^{\frac{x-\mu}{\sqrt{2}\sigma}} e^{-t^2} dt \text{ where } erf(y) = \frac{2}{\sqrt{\pi}} \int_{0}^y e^{-t^2}dt \\
  & = \frac{1}{2} \lim_{y \to \infty} erf(y) + \frac{1}{2} erf \biggr( \frac{x-\mu}{\sqrt{2}\sigma} \biggr) \\
  & = \frac{1}{2} + \frac{1}{2} erf \biggr( \frac{x-\mu}{\sqrt{2}\sigma} \biggr) \text{ By the properties of } erf \\
  & = \frac{1}{2} \biggr[ 1 + erf \biggr( \frac{x-\mu}{\sqrt{2}\sigma} \biggr) \biggr] \\
  \end{split}$$

## Student's t-distribution

Let $Z \sim \mathcal{N}(0, 1)$, $V \sim \chi^2(\nu)$ and $Z$ and $V$ independent random variables then $Y \sim \frac{Z}{\sqrt{\frac{V}{\nu}}} \sim t(\mathcal{\nu})$ where $t$ is the t-distribution with $\nu > 0$ degrees of freedom

So we have that $f(z) = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^2}$ and $f(v) = \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{\nu}{2}}$

Let $f(z,v)$ by the joint probability density function of $Z$ and $V$. Because $Z$ and $V$ are independent random variables we have that:

$$\begin{split}
  f(z,v) & = f(z)f(v) \\
  & = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}z^2} \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{v}{2}} \\ 
  \end{split}$$

We can specify a variable $t = \frac{z}{\sqrt{\frac{v}{\nu}}}$ such that $z = t\sqrt{\frac{v}{\nu}}$ and $dz = \sqrt{\frac{v}{\nu}}dt$. So:

$$\begin{split}
  F_Z(z) & = \int_{-\infty}^z \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}w^2} dw \\
  F_Z(t) & = \int_{-\infty}^t \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}s^2\frac{v}{\nu}} \sqrt{\frac{v}{\nu}}ds \\
  \frac{F_Z(t)}{dt} & = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}t^2\frac{v}{\nu}} \sqrt{\frac{v}{\nu}} \\ 
  f_t(t) & = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}t^2\frac{v}{\nu}} \sqrt{\frac{v}{\nu}} \\ 
  \end{split}$$

Also we have that $f(z,v) = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}z^2} \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{v}{2}}$ for $v > 0$ and $z \in (-\infty, \infty)$. In another case $f(z,v) = 0$. Therefore we have that:

$$\begin{split}
  f(z,v) & = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}z^2} \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{v}{2}} \\
  & = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}t^2\frac{v}{\nu}} \sqrt{\frac{v}{\nu}} \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{v}{2}} \\
  & = \frac{1}{\sqrt{2\pi\nu}}e^{-\frac{1}{2}t^2\frac{v}{\nu}}v^{\frac{1}{2}} \frac{1}{2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu}{2}-1}e^{-\frac{v}{2}} \\
  & = \frac{1}{\sqrt{2\pi\nu}2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu-1}{2}}e^{-\frac{v}{2}(1 + \frac{t^2}{\nu})} \\
  \end{split}$$

Where $-\infty < t < \infty$ and $v > 0$. So with this joint probability density function we can recover $f_T(t)$ using the definition of the margina density:

$$\begin{split}
  f_T(t) & = \int_{0}^{\infty} \frac{1}{\sqrt{2\pi\nu}2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}v^{\frac{\nu-1}{2}}e^{-\frac{v}{2}(1 + \frac{t^2}{\nu})} dv \\
  & = \int_{0}^{\infty} \frac{1}{\sqrt{2\pi\nu}2^{\frac{\nu}{2}}\Gamma(\frac{\nu}{2})}2^{\frac{\nu-1}{2}}w^{\frac{\nu-1}{2}}(1 + \frac{t^2}{\nu})^{-\frac{\nu-1}{2}}e^{-w}2(1 + \frac{t^2}{\nu})^{-1} dw \text{ where } w = \frac{v}{2}(1 + \frac{t^2}{\nu}) \\
  & = \int_{0}^{\infty} \frac{1}{\sqrt{\pi\nu}\Gamma(\frac{\nu}{2})}w^{\frac{\nu-1}{2}}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}}e^{-w} dw \\
  & = \frac{1}{\sqrt{\pi\nu}\Gamma(\frac{\nu}{2})}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}} \int_{0}^{\infty} w^{\frac{\nu-1}{2}}e^{-w} dw \\
  & = \frac{1}{\sqrt{\pi\nu}\Gamma(\frac{\nu}{2})}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}} \int_{0}^{\infty} w^{\frac{\nu+1}{2}-1}e^{-w} dw \\
  & = \frac{\Gamma(\frac{\nu+1}{2})}{\sqrt{\pi\nu}\Gamma(\frac{\nu}{2})}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}}\\
  & = \frac{1}{\sqrt{\nu}\frac{\sqrt{\pi}\Gamma(\frac{\nu}{2})}{\Gamma(\frac{\nu+1}{2})}}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}}\\
  & = \frac{1}{\sqrt{\nu}\frac{\Gamma(\frac{1}{2})\Gamma(\frac{\nu}{2})}{\Gamma(\frac{\nu+1}{2})}}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}}\\
  & = \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})}(1 + \frac{t^2}{\nu})^{-\frac{\nu+1}{2}} \\
  \end{split}$$

```{r}
#| echo: false
#| fig-align: center
ggplot() + 
  geom_function(fun=dt, args=list(df=5, ncp=0),
                xlim=c(-5,5),
                color='#2C3E50') + 
  scale_x_continuous(breaks = seq.int(from = -5, to = 5, by = 1)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Student distribution function',
       subtitle = TeX(r'($\nu=5$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

### Cumulative distribution function[^3]

[^3]: You can check out the derivantion in https://statproofbook.github.io/P/f-pdf

$$\begin{split}
  F_{\nu}(t) & = \int_{-\infty}^t \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})}(1 + \frac{s^2}{\nu})^{-\frac{\nu+1}{2}} ds \\
  & = \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})} \int_{-\infty}^t (1 + \frac{s^2}{\nu})^{-\frac{\nu+1}{2}} ds \\
  & = \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})} \int_{-\infty}^t (\frac{\nu + s^2}{\nu})^{-\frac{\nu+1}{2}} ds \\
  & = \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})} \int_{0}^{\frac{\nu}{\nu + t^2}} (\frac{1}{x})^{-\frac{\nu+1}{2}} \frac{\nu}{2x^2\sqrt{\nu(\frac{1-x}{x})}}dx \text{ where } x = \frac{\nu}{\nu + s^2} \text{ and } s < 0 \\
  & = \frac{1}{2}\frac{1}{\beta(\frac{1}{2},\frac{\nu}{2})} \int_{0}^{\frac{\nu}{\nu + t^2}} x^{\frac{\nu}{2}-1}(1-x)^{-\frac{1}{2}} dx \\
  & = \frac{1}{2}\frac{1}{\beta(\frac{1}{2},\frac{\nu}{2})} \int_{0}^{\frac{\nu}{\nu + t^2}} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1} dx \\
  & = \frac{1}{2}\frac{\beta(\frac{\nu}{\nu + t^2};\frac{\nu}{2},\frac{1}{2})}{\beta(\frac{1}{2},\frac{\nu}{2})} \\
  & = \frac{1}{2}\frac{\beta(\frac{\nu}{\nu + t^2};\frac{\nu}{2},\frac{1}{2})}{\beta(\frac{\nu}{2},\frac{1}{2})} \\
  & = \frac{1}{2}I_{\frac{\nu}{\nu + t^2}}(\frac{\nu}{2},\frac{1}{2})
  \end{split}$$

In the case of $t \geq 0$ check out [@johnson_continuous_1995, page 364]:

$$\begin{split}
  F_{\nu}(t) & = F_{\nu}(0) + F_{\nu}(t) -  F_{\nu}(0)\\
  & = \frac{1}{2} + \int_0^{t} \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})}(1 + \frac{s^2}{\nu})^{-\frac{\nu+1}{2}} ds \\
  & = \frac{1}{2} + \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})} \int_{1}^{\frac{\nu}{\nu + t^2}} (\frac{1}{x})^{-\frac{\nu+1}{2}} \frac{-\nu}{2x^2\sqrt{\nu(\frac{1-x}{x})}}dx \text{ where } x = \frac{\nu}{\nu + s^2} \text{ and } s \geq 0 \\
  & = \frac{1}{2} - \frac{1}{\sqrt{\nu}\beta(\frac{1}{2},\frac{\nu}{2})} \int_{\frac{\nu}{\nu + t^2}}^{1} (\frac{1}{x})^{-\frac{\nu+1}{2}} \frac{-\nu}{2x^2\sqrt{\nu(\frac{1-x}{x})}}dx \\
  & = \frac{1}{2} + \frac{1}{2}\frac{1}{\beta(\frac{1}{2},\frac{\nu}{2})} \int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{-\frac{1}{2}} dx \\
  & = \frac{1}{2} + \frac{1}{2}\frac{1}{\beta(\frac{\nu}{2},\frac{1}{2})} \int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1} dx \\
  \end{split}$$

We have the following result:

$$\begin{split}
 \frac{\int_{0}^{\frac{\nu}{\nu + t^2}} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1} + \int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1}}{\beta(\frac{\nu}{2},\frac{1}{2})} & = 1 \\
 \frac{\int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1}}{\beta(\frac{\nu}{2},\frac{1}{2})} & = 1  - \frac{\int_{0}^{\frac{\nu}{\nu + t^2}} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1}}{\beta(\frac{\nu}{2},\frac{1}{2})} \\
 \frac{\int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1}}{\beta(\frac{\nu}{2},\frac{1}{2})} & = 1  - I_{\frac{\nu}{\nu + t^2}}(\frac{\nu}{2},\frac{1}{2}) \\
  \end{split}$$

Using the above result we have the following:

$$\begin{split}
  F_{\nu}(t) & = \frac{1}{2} + \frac{1}{2}\frac{1}{\beta(\frac{\nu}{2},\frac{1}{2})} \int_{\frac{\nu}{\nu + t^2}}^{1} x^{\frac{\nu}{2}-1}(1-x)^{\frac{1}{2}-1} dx \\
 & = \frac{1}{2} + \frac{1}{2}\biggr[ 1  - I_{\frac{\nu}{\nu + t^2}}(\frac{\nu}{2},\frac{1}{2}) \biggr] \\
 & = 1 - \frac{1}{2}I_{\frac{\nu}{\nu + t^2}}(\frac{\nu}{2},\frac{1}{2}) \text{ where } t \geq 0
\end{split}$$

## Transformation of random variables

### Transformations of a single random variable

Assume that $X$ is a random variable where $X$ is real-valued. In that sense $X: \Omega \longrightarrow \mathbb{R}$ where $\Omega$ is the sample space. Now let $g: \mathbb{R} \longrightarrow \mathbb{R}$ be a continuous function. Then we can define a random variable $Y = g(X)$ where the idea is to find the cumulative distribution function $F_Y(y)$ given the cumulative distribution function $F_X(y)$

The classical example is the case where $X \sim \mathcal{N}(0,1)$ where we can do find $Y = X^2$. We can proceed in the following way:

$$\begin{split}
  F_Y(y) & = Pr[Y \leq y] \\
  & = Pr[X^2 \leq y] \\
  & = Pr[|X| \leq \sqrt{y}] \\
  & = Pr[ -\sqrt{y} \leq X \leq \sqrt{y}] \\
  & = \int_{-\sqrt{y}}^{\sqrt{y}} \frac{1}{\sqrt{2\pi}}e^{\frac{1}{2}x^2} dx \\
  & = 2 \int_{0}^{\sqrt{y}} \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}x^2}
  dx \\
  & = 2 \int_{0}^y \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}u}
  \frac{1}{2\sqrt{u}} du \text{ where } u = x^2 \text{ with } u >
  0 \\
  & = \int_{0}^y \frac{1}{\sqrt{2u\pi}}e^{-\frac{1}{2}u} du \\
  \frac{dF_Y(y)}{dy} & = \int_{0}^y
  \frac{1}{\sqrt{2u\pi}}e^{-\frac{1}{2}u} du \\
  f_Y(y) & = \frac{1}{\sqrt{2y\pi}}e^{-\frac{1}{2}y} \\
  & = \frac{1}{2^{\frac{1}{2}}\sqrt{\pi}}y^{-\frac{1}{2}}e^{-\frac{1}{2}y} \\
  & = \frac{1}{2^{\frac{1}{2}}\Gamma(\frac{1}{2})}y^{\frac{1}{2}-1}e^{-\frac{y}{2}} \text{ where } y > 0 \\
  & = \chi^2_1
  \end{split}$$

Now assume $g$ is strictly increasing, $x_1 < x_2 \Longrightarrow g(x_1) < g(x_2)$, and differentiable function.

We have that $g$ is a one-to-one function and its inverse $g^{-1}$ is also a function and one-to-one. Furthermore, $g^{-1}$ is also strictly increasing

$$\begin{split}
  F_Y(y) & = Pr[Y < y] \\ 
  & = Pr[g(X) < y] \\
  & = Pr[X < g^{-1}(y)] \\
  & = \int_{-\infty}^{g^{-1}(y)} f_X(x) dx \\
  & = \int_{-\infty}^{y} f_X(g^{-1}(t)) (g^{-1})'(t)dt \text{ where } x = g^{-1}(t) \\
  & = \int_{-\infty}^{y} f_X(g^{-1}(t)) \frac{1}{g'(g^{-1}(t))}dt \text{ where } g'(g^{-1}(t)) \neq 0 \\
  \frac{dF_Y(y)}{dy} & = f_X(g^{-1}(y)) \frac{1}{g'(g^{-1}(y))} \\
  f_Y(y) & = f_X(g^{-1}(y)) \frac{1}{g'(g^{-1}(y))}
  \end{split}$$

We can check the same result for $g$ when is strictly decreasing, $x_1 < x_2 \Longrightarrow g(x_1) > g(x_2)$, and differentiable function.

We have that $g$ is a one-to-one function and its inverse $g^{-1}$ is also a function and one-to-one. Furthermore, $g^{-1}$ is also strictly decreasing

$$\begin{split}
  F_Y(y) & = Pr[Y < y] \\ 
  & = Pr[g(X) < y] \\
  & = Pr[X > g^{-1}(y)] \\
  & = \int_{g^{-1}(y)}^{\infty} f_X(x) dx \\
  & = \int_{y}^{\infty} f_X(g^{-1}(t)) (g^{-1})'(t)dt \text{ where } x = g^{-1}(t) \\
  & = \int_{y}^{\infty} f_X(g^{-1}(t)) \frac{1}{g'(g^{-1}(t))}dt \text{ where } g'(g^{-1}(t)) \neq 0 \\
  & = 1 -  \int_{-\infty}^{y} f_X(g^{-1}(t)) \frac{1}{g'(g^{-1}(t))}dt \\
  \frac{dF_Y(y)}{dy} & = f_X(g^{-1}(y)) \frac{-1}{g'(g^{-1}(y))} \\
  f_Y(y) & = f_X(g^{-1}(y)) \frac{-1}{g'(g^{-1}(y))}
  \end{split}$$

Taking into account both results we have that:

$$f_Y(y) = f_X(g^{-1}(y)) \frac{1}{|g'(g^{-1}(y))|}$$

We can apply this result to the following case where $X \sim \mathcal{N}(\mu, \sigma^2)$ and the idea is to find the distribution of $Y = e^X$

-   $\frac{de^x}{dx} = e^x$
-   $g^{-1}(y) = log_ey$
-   $\frac{de^x}{dx} = e^x \Big|_{x=log_e y} = y > 0$
-   $f_Y(y) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{log_ey - \mu)}{\sigma}} \frac{1}{y} = \frac{1}{y\sigma\sqrt{2\pi}}e^{-\frac{1}{2}(\frac{log_ey - \mu)}{\sigma}}$

Now we can generalize for $X_1$ and $X_2$ to find the probability density function using the concept of **Jacobian**

Let $f_{X_1,X_2}(x_1,x_2)$ the probability density function, $Y_1 = g_1(X_1, X_2)$ and $Y_2 = g_2(X_1, X_2)$ then we have:

-   $y_1 = g_1(x_1, x_2)$, $y_2 = g_2(x_1, x_2)$

-   We need to solve for $x_1$ and $x_2$ in terms of $y_1$ and $y_2$

    -   $x_1 = h(y_1, y_2)$
    -   $x_2 = h(y_1, y_2)$

-   We need to compute the following Jacobin:

$$\mathbf{J} = \begin{bmatrix}
                \frac{dx_1}{dy_1} & \frac{dx_1}{dy_2} \\
                \frac{dx_2}{dy_1} & \frac{dx_2}{dy_2} 
                \end{bmatrix}$$

-   Then we need to calculate $|det(\mathbf{J})|$, that is the absolute value of the determinant of the Jacobian, $\bigg| \frac{dx_1}{dy_1}\frac{dx_2}{dy_2} - \frac{dx_1}{dy_2}\frac{dx_2}{dy_1} \bigg|$

-   Finally we have that $f_{Y_1, Y_2}(y_1,y_2) = f_{X_1, X_2}(h(y_1, y_2),h(y_1, y_2))|det(\mathbf{J})|$

## F-distribution

If $X_1$ and $X_2$ are independent random variables distributed as $\chi^2_{\nu_1}$ and $\chi^2_{\nu_2}$ then the distribution of $\frac{\frac{X_1}{\nu_1}}{\frac{X_2}{\nu_2}}$ is the F-distribution with $\nu_1$ and $\nu_2$ degrees of freedom.

```{r}
#| echo: false
#| fig-align: center
ggplot() + 
  geom_function(fun=df, args=list(df1=5, df2=4),
                xlim=c(0,10),
                color='#2C3E50') + 
  scale_x_continuous(breaks = seq.int(from = 0, to = 10, by = 1)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'F distribution function',
       subtitle = TeX(r'($\nu_1=5,\nu_2=4$)')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

### Probability density function [^4]

[^4]: Check this to specify the null hypothesis and calculate the expected values <https://online.stat.psu.edu/stat500/lesson/8/8.1>

-   Let $F = \frac{\frac{X_1}{\nu_1}}{\frac{X_2}{\nu_2}}$ and $W = Y$

-   We have that $f = \frac{\frac{x_1}{\nu_1}}{\frac{x_2}{\nu_2}}$ and $w = y$. So $x_1 = fw\frac{\nu_1}{\nu_2}$ and $x_2 = w$

-   The Jacobian is defined as:

$$\mathbf{J} = \begin{bmatrix}
                w\frac{\nu_1}{\nu_2} & f\frac{\nu_1}{\nu_2} \\
                0 & 1 
                \end{bmatrix}$$

-   We have that $|det(\mathbf{J})| = |w\frac{\nu_1}{\nu_2}| = w\frac{\nu_1}{\nu_2}$ because $w$ is a realization of a random variable that it is distributed as a $\chi^2_{\nu_2}$, $\nu_1 > 0$ and $\nu_2 > 0$

-   The we have that:

$$\begin{split}
  f_{F,W}(f,w) & = f_{X_1,X_2}(fw\frac{\nu_1}{\nu_2},w)|det(\mathbf{J})| \\
  & = f_{X_1}(fw\frac{\nu_1}{\nu_2})f_{X_2}(w)|det(\mathbf{J})| \\
  & = \frac{1}{2^{\frac{\nu_1}{2}}\Gamma(\frac{\nu_1}{2})}(fw\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}-1}e^{-fw\frac{\nu_1}{2\nu_2}} \frac{1}{2^{\frac{\nu_2}{2}}\Gamma(\frac{\nu_2}{2})}w^{\frac{\nu_2}{2}-1}e^{-\frac{w}{2}} w\frac{\nu_1}{\nu_2} \\
  & = \frac{(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}}{2^{\frac{\nu_1+ \nu_2}{2}}\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})}w^{\frac{\nu_1+\nu_2}{2}-1}e^{-\frac{w}{2}(f\frac{\nu_1}{\nu_2} + 1)} \\
  \end{split}$$

Now we need to recover $f_F(f)$ using the probability marginal density function:

$$\begin{split}
  f_F(f) & = \int_0^{\infty} f_{F,W}(f,w) dw \\
  & = \int_0^{\infty} \frac{(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}}{2^{\frac{\nu_1+ \nu_2}{2}}\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})}w^{\frac{\nu_1+\nu_2}{2}-1}e^{-\frac{w}{2}(f\frac{\nu_1}{\nu_2} + 1)} dw \\
  & = \frac{(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}}{2^{\frac{\nu_1+ \nu_2}{2}}\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})} \int_0^{\infty} w^{\frac{\nu_1+\nu_2}{2}-1}e^{-\frac{w}{2}(f\frac{\nu_1}{\nu_2} + 1)} dw \\ 
  & = \frac{(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}}{2^{\frac{\nu_1+ \nu_2}{2}}\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})} \frac{\Gamma(\frac{\nu_1 + \nu_2}{2})}{(\frac{1}{2}(f\frac{\nu_1}{\nu_2}+1))^{\frac{\nu_1 + \nu_2}{2}}} \int_0^{\infty} \frac{w^{\frac{\nu_1+\nu_2}{2}-1}e^{-\frac{w}{2}(f\frac{\nu_1}{\nu_2} + 1)}(f\frac{\nu_1}{\nu_2}+1))^{\frac{\nu_1 + \nu_2}{2}}}{\Gamma(\frac{\nu_1 + \nu_2}{2})} dw \\ 
  & = \frac{(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}}{2^{\frac{\nu_1+ \nu_2}{2}}\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})} \frac{\Gamma(\frac{\nu_1 + \nu_2}{2})}{(\frac{1}{2}(f\frac{\nu_1}{\nu_2}+1))^{\frac{\nu_1 + \nu_2}{2}}}  \text{ where } \int_0^{\infty} \frac{w^{\frac{\nu_1+\nu_2}{2}-1}e^{-\frac{w}{2}(f\frac{\nu_1}{\nu_2} + 1)}(f\frac{\nu_1}{\nu_2}+1))^{\frac{\nu_1 + \nu_2}{2}}}{\Gamma(\frac{\nu_1 + \nu_2}{2})} dw = 1 \\
  & = \frac{\Gamma(\frac{\nu_1 + \nu_2}{2})}{\Gamma(\frac{\nu_1}{2})\Gamma(\frac{\nu_2}{2})} (\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1} (f\frac{\nu_1}{\nu_2}+1)^{-\frac{\nu_1 + \nu_2}{2}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2},\frac{\nu_2}{2})} (\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1} (f\frac{\nu_1}{\nu_2}+1)^{-\frac{\nu_1 + \nu_2}{2}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2},\frac{\nu_2}{2})} (\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1} (f\frac{\nu_1}{\nu_2}+1)^{-\frac{\nu_1}{2}-\frac{\nu_2}{2}}
  \end{split}$$

### Cumulative distribution function

According to [@johnson_continuous_1995, page 248] and [@johnson_continuous_1995, page 325] we have the following result:

$$F = \frac{\nu_2}{\nu_1}\frac{X}{1-X} \sim f(\nu_1, \nu_2) \text{ where } X \sim \beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})$$ That is $F$ is distributed as a F-distribution with $\nu_1$ and $\nu_2$ as parameters

We have that $g(x) = \frac{\nu_2}{\nu_1}\frac{x}{1-x}$ is an increasing an continuos function in $x \in (0,1)$ taking into account that $\nu_2 > 0$ and $\nu_1 > 0$. Therefore we have that:

-   $g^{-1}(f) = \frac{f\nu_1}{\nu_2 + f\nu_1}$
-   $g'(f) = \frac{\nu_2}{\nu_1(1-f)^2}$
-   $\frac{1}{g'(g^{-1}(f))} = \frac{\frac{1}{1}}{\frac{\frac{\nu_2}{1}}{\nu_1(1-\frac{f\nu_1}{\nu_2 + f\nu_1})^2}} = \frac{\frac{1}{1}}{\frac{\frac{\nu_2}{1}}{\nu_1(\frac{\nu_2}{\nu_2 + f\nu_1})^2}} = \frac{\nu_1\nu_2}{(\nu_2 + f\nu_1)^2}$

So we have the following result:

$$\begin{split}
  f_F & = f_X(g^{-1}(f)) \frac{1}{g'(g^{-1}(f))} \\ 
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{f\nu_1}{\nu_2 + f\nu_1})^{\frac{\nu_1}{2}-1}(1-\frac{f\nu_1}{\nu_2 + f\nu_1})^{\frac{\nu_2}{2} - 1} \frac{1}{\nu_1\nu_2}(\frac{\nu_2 + f\nu_1}{\nu_2})^2 \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{f\nu_1}{\nu_2 + f\nu_1})^{\frac{\nu_1}{2}-1} (\frac{\nu_2}{\nu_2 + f\nu_1})^{\frac{\nu_2}{2} - 1} \frac{\nu_1\nu_2}{(\nu_2 + f\nu_1)^2} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}\nu_1^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}\nu_2^{\frac{\nu_2}{2}}\frac{1}{(\nu_2 + f\nu_1)^{\frac{\nu_1}{2} + \frac{\nu_2}{2}}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}\nu_2^{\frac{\nu_2}{2}}\nu_2^{\frac{\nu_1}{2}}\frac{1}{(\nu_2 + f\nu_1)^{\frac{\nu_1}{2} + \frac{\nu_2}{2}}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}\nu_2^{\frac{\nu_1}{2}+\frac{\nu_2}{2}}\frac{1}{(\nu_2 + f\nu_1)^{\frac{\nu_1}{2} + \frac{\nu_2}{2}}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}\frac{1}{(\frac{\nu_2}{\nu_2} + \frac{f\nu_1}{\nu_2})^{\frac{\nu_1}{2} + \frac{\nu_2}{2}}} \\
  & = \frac{1}{\beta(\frac{\nu_1}{2}, \frac{\nu_2}{2})}(\frac{\nu_1}{\nu_2})^{\frac{\nu_1}{2}}f^{\frac{\nu_1}{2}-1}(1 + \frac{f\nu_1}{\nu_2})^{-\frac{\nu_1}{2} - \frac{\nu_2}{2}} \\
  \end{split}$$

We can use the above result to find the cumulative function in the following way:

$$\begin{split}
  F_F(f) & = Pr[F \leq f] \\
  & = Pr[\frac{\nu_2}{\nu_1}\frac{X}{1-X} \leq f]
  \end{split}$$

# Testing Group Frequencies: chisq.test()

## Toy data

### Base R

```{r}
#| echo: true
data_chisq_test <- rep(c(1:4), times = c(25,25,25,20))
data_chisq_test
tmp.tab <- table(data_chisq_test)
tmp.tab
```

### Tidyverse

```{r}
#| echo: true
tmp_tab <- tibble(x =data_chisq_test)|> 
  count(x, name = 'observed') |> 
  mutate(expected = (1/4)*sum(observed))
tmp_tab
```

## chisq.test()

### By hand

#### Chi-squared distribution: $\chi^2$

$\sum_{i=1}^k Z_i^2 \sim \chi^2(k)$ where $Z_i \sim \mathcal{N}(0,1)$ and $Z_1, \ldots, Z_k$ are independent random variables

#### Derivation of the chi-squared probability density function

$$\begin{split}
  M_{\sum_{i=1}^k Z_i^2}(t) & = E[e^{t\sum_{i=1}^k Z_i^2}] \\
  & = E[e^{t(Z_1^2 + \cdots + Z_k^2)}] \\
  & = E[e^{tZ_1^2} \cdots e^{tZ_k^2}] \\
  & = E[e^{tZ_1^2}] \cdots E[e^{tZ_k^2}] \text{ Because } Z_1, \ldots Z_k \text{ are independent random variables} \\
  & = \frac{1}{(1 - 2t)^{\frac{1}{2}}} \cdots \frac{1}{(1 - 2t)^{\frac{1}{2}}} \text{ Because } Z_i^2 \sim \chi^2(1) \\
  &  = \frac{1}{(1 - 2t)^{\frac{k}{2}}}
  \end{split}$$

So $\sum_{i=1}^k Z_i^2$ and $\chi^2(k)$ has the same moment generating function. According to [@rice_mathematical_2021, Chapter 4, page 155] the moment-generating function uniquely determines the probability density function under certain conditions. Using this we can conclude that$\sum_{i=1}^k Z_i^2 \sim \chi^2(k)$

```{r}
#| echo: false
#| fig-align: center
ggplot() +
  geom_function(aes(color="Df: 1"), 
                fun = dchisq, args=list(df=1), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 2"), 
                fun = dchisq, args=list(df=2), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 3"), 
                fun = dchisq, args=list(df=3), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 4"), 
                fun = dchisq, args=list(df=4), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 5"), 
                fun = dchisq, args=list(df=5), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 6"), 
                fun = dchisq, args=list(df=6), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 7"), 
                fun = dchisq, args=list(df=7), 
                xlim = c(0,15)) +
  geom_function(aes(color="Df: 8"), 
                fun = dchisq, args=list(df=8), 
                xlim = c(0,15)) +
  scale_color_tq() +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = 'Chi-squared distribution',
       subtitle = 'Df: Degrees of freedom') +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

### Test for given probabilities

$H_0: p_1 = \frac{1}{4} \land p_2 = \frac{1}{4} \land p_3 = \frac{1}{4} \land p_4 = \frac{1}{4}$

$H_1: p_1 \neq \frac{1}{4} \lor p_2 \neq \frac{1}{4} \lor p_3 = \frac{1}{4} \lor p_4 \neq \frac{1}{4}$

$\chi^2 = \sum_{i=1}^n \frac{(Observed_i - Expected_i)^2}{Expected_i} = \frac{25 - 95\frac{1}{4}}{95\frac{1}{4}} + \frac{25 - 95\frac{1}{4}}{95\frac{1}{4}} + \frac{25 - 95\frac{1}{4}}{95\frac{1}{4}} + \frac{20 - 95\frac{1}{4}}{95\frac{1}{4}}$

```{r}
#| echo: true
chi_square_statistic <- ((tmp_tab$observed - tmp_tab$expected)^2 / 
                           tmp_tab$expected) |> 
  sum()
chi_square_statistic
degrees_of_freedom <- nrow(tmp_tab) - 1
degrees_of_freedom
p_value <- pchisq(chi_square_statistic, 
                  df = degrees_of_freedom,
                  lower.tail = FALSE)
p_value
```

$\text{p-value} = \mathbf{P}(\chi^2(3) > 0.7894737) = 0.8519831$

Data shows no evidence that the groups in the population are of unequal size, under the assumption of random sampling. In general, a $\text{p-value}$ less than 0.10, 0.05 or 0.01 suggests that there is a difference between groups

#### Base R

```{r}
#| echo: true
chisq_test <- chisq.test(x = table(data_chisq_test), 
                         p = rep(1/length(tmp.tab),
                                 length(tmp.tab)))
str(chisq_test)
chisq_test$statistic
chisq_test$parameter
chisq_test$p.value
chisq_test$method
chisq_test$observed
chisq_test$expected
```

#### Tidymodels

```{r}
#| echo: true
tibble(x = data_chisq_test) |>
  mutate(x = factor(x, ordered = FALSE)) |> 
  chisq_test(response = x,
             p = c(1/4, 1/4, 1/4, 1/4))
```

### Test for independence[^5]

```{r}
#| echo: true
#| message: false
segmentation <- read_csv(file = '000_data_sets/006_rintro-chapter5.csv')
segmentation |> 
  count(subscribe, ownHome) |> 
  pivot_wider(id_cols = subscribe,
              names_from = ownHome,
              values_from = n)
```

$H_0: p_{11} = \frac{260}{300}\frac{159}{300} \land p_{12} = \frac{260}{300}\frac{141}{300} \land p_{21} = \frac{40}{300}\frac{159}{300} \land p_{22} = \frac{40}{300}\frac{141}{300}$

$H_1: p_{11} \neq \frac{260}{300}\frac{159}{300} \lor p_{12} \neq \frac{260}{300}\frac{141}{300} \lor p_{21} \neq \frac{40}{300}\frac{159}{300} \lor p_{22} \neq \frac{40}{300}\frac{141}{300}$

$\chi^2 = \sum_{i=1}^n \frac{(Observed_i - Expected_i)^2}{Expected_i} = \frac{(137 - 300\frac{260}{300}\frac{159}{300})^2}{300\frac{260}{300}\frac{159}{300}} + \frac{(123 - 300\frac{260}{300}\frac{141}{300})^2}{300\frac{260}{300}\frac{141}{300}} + \frac{(22 - 300\frac{40}{300}\frac{159}{300})^2}{300\frac{40}{300}\frac{159}{300}} + \frac{(18 - 300\frac{40}{300}\frac{141}{300})^2}{300\frac{40}{300}\frac{141}{300}}$

[^5]: See this resources to understand the calculation of observed and expected values [https://online.stat.psu.edu/stat500/lesson/8/8.1](https://online.stat.psu.edu/stat500/lesson/8/8.1)

# Testing Observed Proportions: binom.test()

## Toy data

### Base R

```{r}
#| echo: true
table(segmentation$gender)
```

### Tidyverse

```{r}
#| echo: true
segmentation |> 
  count(gender)
```

In our case we define as a success that $gender = Female$ which mean the $Female$ is coded us $1$

## binom.test()

### By hand

$H_0: p = 0.5$

$H_1: p \neq 0.5$

$B = \sum_{i=1}^n x_i = 157$ where $x_i \in {0,1}$

$\hat{p} = \frac{157}{300} \approx 0.5233$ and $1- \hat{p} = 1 - \frac{157}{300} \approx 0.4766$

$\text{p-value} = \mathbf{P}(X \leq 143) + \mathbf{P}(X \geq 157) \approx 0.2264879 + 0.2264879 \approx 0.4529757$

Using the help of R without using a predefined table we have that:

```{r}
#| echo: true
pbinom(q = 300 - 157, 
       size = 300, 
       prob = 0.5, 
       lower.tail = TRUE) + 
pbinom(q = 157 - 1, 
       size = 300, 
       prob = 0.5, 
       lower.tail = FALSE)
```

Based on the data at hand we don't have sufficient evidence to reject that $p=0.5$

In the case of confidence interval for $\alpha = 0.05$ we have the following result:

$$\begin{split}
  I_x^{-1} \biggr(p = \frac{0.05}{2}; a = 157, b = 300 - 157 + 1\biggr) & < p < I_x^{-1}\biggr(p = 1 - \frac{0.05}{2}; a = 157 + 1, b = 300 - 157\biggr) \\
  0.4651595 & < p < 0.58104184 \\
  \end{split}$$ Where $I_x^{-1}(p; a, b)$ is the beta quantile function taking into account that $I_x(a,b)$ is the beta cumulative distribution function. Therefore we have:

$$\begin{split}
  I_x(a,b) & = p \\
  x & = I_x^{-1}(p; a,b)
  \end{split}$$

The above interval means that in the long run $95\%$ of confidence intervals constructed in this manner will contain the true parameter

Also this confidence interval is obtained in the following way by taking into account that the interval contains all values of $p$ that are not rejected by the test at confidence level $\alpha$:

```{=tex}
\begin{align*}
 P_{p_L}(X \geq x) & = \sum_{k = x}^n \binom{n}{k}
 p_L^k(1-p_L)^{n-k} & = \frac{\alpha}{2} \\
 & = I_{p_L}(x, n - x + 1) & = \frac{\alpha}{2} \\
 & = p_L & = I_{p_L}^{-1}(\frac{\alpha}{2}; x, n - x + 1) \\
 \end{align*}
```
```{=tex}
\begin{align*}
 P_{p_U}(X \leq x) & = 1 -  P_{p_U}(X \geq x + 1) & = \frac{\alpha}{2} \\
 & = 1 - I_{p_U}(x + 1, n - (x + 1) + 1) & = \frac{\alpha}{2} \\
 & = 1 - I_{p_U}(x + 1, n - x) & = \frac{\alpha}{2} \\
 & = I_{p_U}(x + 1, n - x) & = 1 - \frac{\alpha}{2} \\
 & = p_U & = I_{p_U}^{-1}(1 - \frac{\alpha}{2}; x + 1, n - x) \\
 \end{align*}
```
### Base R

```{r}
#| echo: true
binom_test <- binom.test(x = 157, n = 300, p = 0.5, 
           alternative = "two.sided", conf.level = 0.95)
str(binom_test)
binom_test$statistic
binom_test$estimate
binom_test$p.value
binom_test$conf.int
```

### Tidyverse

```{r}
binom_test |> 
  tidy()
```

# Testing Group Means: t.test()

## Toy data

### Tidyverse

```{r}
segmentation |> 
  ggplot() + 
  geom_histogram(aes(x = income),
                 color='black') +
  facet_wrap(facets = vars(ownHome))
```

```{r}
sample_statistics <- segmentation |> 
  group_by(ownHome) |> 
  summarise(mean_income = mean(income),
                         # sample variance
                         ## use n - 1 in the denominator
            var_income = var(income),
            n = n())
sample_statistics
```

## t.test()

### By hand

$H_0: \mu_{ownNo} - \mu_{ownYes}= 0$

$H_1: \mu_{ownNo} - \mu_{ownYes} \neq 0$

$t = \frac{\overline{ownNo} - \overline{ownYes}}{\sqrt{\frac{s_{ownNo}^2}{n_{ownNo}} + \frac{s_{ownYes}^2}{n_{ownYes}}}} = \frac{47391.01 - 54934.68}{\sqrt{ \frac{358692875}{159} + \frac{430890091}{141}}} \approx -3.273094$

-   Degrees of freedom

$\nu \approx \frac{(\frac{s_1^2}{N_1} + \frac{s_2^2}{N_2})^2}{\frac{(\frac{s_1^2}{N_1})^2}{N_1-1} + \frac{(\frac{s_2^2}{N_2})^2}{N_2-1}} = \frac{(\frac{358692875}{159} + \frac{430890091}{141})^2}{\frac{(\frac{358692875}{159})^2}{159-1} + \frac{(\frac{430890091}{141})^2}{141-1}} = 285.2521$

Using R we have the following result:

```{r}
((sample_statistics$var_income[1] / sample_statistics$n[1]) +
   (sample_statistics$var_income[2] / sample_statistics$n[2]))^2 /
  ((sample_statistics$var_income[1] / sample_statistics$n[1])^2 /
     (sample_statistics$n[1] - 1) + (sample_statistics$var_income[2] / 
     sample_statistics$n[2])^2 /
     (sample_statistics$n[2] - 1))
```

$\text{p-value} = \mathbf{P}(T \leq t) + \mathbf{P}(T \geq t) \approx 0.0005973553 + 0.0005973553 \approx 0.001194711$

-   Confidence interval

$$P \Biggr( t_L < \frac{\overline{x}_{ownNo} - \overline{x}_{ownYes} - (\mu_{ownNo} - \mu_{ownYes})}{\sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } +\frac{s^2_{ownYes}}{n_{ownYes}}}} < t_U \Biggr) = 0.95$$ We need to specify $t_L$ and $t_U$:

```{r}
#| echo: true
t_L <- qt(p = 0.025, df = 285.25, lower.tail = TRUE)
t_L
t_U <- qt(p = 0.975, df = 285.25, lower.tail = TRUE)
t_U
```

Therefore we have that:

$$P \Biggr( -t_{0.025,\nu} < \frac{\overline{x}_{ownNo} - \overline{x}_{ownYes} - (\mu_{ownNo} - \mu_{ownYes})}{\sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } +\frac{s^2_{ownYes}}{n_{ownYes}}}} < t_{0.025,\nu} \Biggr) = 0.95$$ Where $\nu \approx \frac{(\frac{s_{ownNo}^2}{n_{ownNo}} + \frac{s_2^2}{n_{ownYes}})^2}{\frac{(\frac{s_{ownNo}^2}{n_{ownNo}})^2}{n_{ownNo}-1} + \frac{(\frac{s_2^2}{n_{ownYes}})^2}{n_{ownYes}-1}} = 285.2521$

$$P(-7543.674 - 1.968315\times2304.753 < \mu_{ownNo} - \mu_{ownYes} < -7543.674 - 1.968315\times2304.753) = 0.95$$

$$P(-12080.16 < \mu_{ownNo} - \mu_{ownYes} < -3007.193) = 0.95$$

-   In the long run 95% of confidence intervals constructed in this manner will contain the true parameter

### Base R

```{r}
#| echo: true
t_test <- t.test(income ~ ownHome, data = segmentation,
                 alternative ='two.sided', mu = 0,
                 # See https://www.statology.org/paired-vs-unpaired-t-test/
                 paired = FALSE,
                 # Welch's t-test
                 var.equal = FALSE,
                 conf.level = 0.95)
str(t_test)
```

### Tidyverse

```{r}
#| echo: true
segmentation |> 
  t_test(formula = income ~ ownHome,
         alternative = "two-sided",
         order = c("ownNo", "ownYes"),
         mu = 0,
         conf_level = 0.95)
```

# Testing Multiple Group Means: aov() and anova()

## Toy data

### Tidyverse

```{r}
#| echo: true
toy_data <- tribble(
  ~TimetoFail,  ~Temp, ~Brand, ~Group,
         181L, "Room",    "A",   "RA",
         187L, "Room",    "B",   "RB",
         150L, "Room",    "C",   "RC",
         173L, "Room",    "D",   "RD",
          85L, "Cold",    "A",   "CA",
          80L, "Cold",    "B",   "CB",
          93L, "Cold",    "C",   "CC",
          87L, "Cold",    "D",   "CD",
         180L, "Room",    "A",   "RA",
         192L, "Room",    "B",   "RB",
         159L, "Room",    "C",   "RC",
         190L, "Room",    "D",   "RD",
          85L, "Cold",    "A",   "CA",
          87L, "Cold",    "B",   "CB",
         100L, "Cold",    "C",   "CC",
          98L, "Cold",    "D",   "CD"
  )
toy_data |> 
  arrange(Temp, Brand)
```

```{r}
#| echo: true
toy_data |> 
  ggplot(aes(x = Brand, y = TimetoFail, fill = Temp)) + 
  geom_point(position = position_jitter(width = 0.1),
             shape=21, color='black')
```

## One way Anova: aov() & anova()

In the case of one way anova it is assumed that there are many $y_{ij}$ with $i = 1, \ldots, n_{j}$ and $j = 1, \ldots, a$ where $n = \sum_{j = 1}^a n_j$

For example in the case of `toy_data` with the variables `TimetoFail` and `Temp` we have that $j = 1, 2$, $n_1 = 8$, $n_2 = 8$:

```{r}
#| echo: true
toy_data |> 
  group_by(Temp) |> 
  summarize(n = n())
```

Also it is assumed that each $y_{ij}$ is a realization of a random variable:

$$Y_{ij} \sim \mathcal{N}(\mu_{j}, \sigma^2)$$ Specifically it is assumed the following:

$$\begin{split}
  Y_{ij} & = \mu_j + \epsilon_{ij} \\
  & = \mu + \beta_j + \epsilon_{ij}
  \end{split}$$

Where $\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$ and $\mu_i = \mu + \beta_j$. In the case of the variables `TimetoFail` and `Temp` we need to estimate the following parameters: $\mu_1, \mu_2, \mu, \sigma^2$. Therefore we have $a + 1$ parameters to estimate because $\sigma^2$ can be estimated using to estimations of $\mu_1, \ldots, \mu_a, \mu$.

In `R` the following constrain are imposed using `contr.treatment`:

$$\beta_1 = 0 \text{ and } \mu = \mu_1$$

The problem to solve is the following:

$$\hat{\mu}, \hat{\beta_j} = \arg \min_{\mu, \beta_j} \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \mu - \beta_j)^2$$ In the literature the following notation is used:

-   $y_{\cdot j} = \sum_{i=1}^{n_j} y_{ij} \text{ sum of group } j$
-   $y_{\cdot\cdot} = \sum_{j=1}^a \sum_{i=1}^{n_j} y_{ij} \text{ sum of all observations}$
-   $\overline{y}_{\cdot j} = \frac{1}{n_j} \sum_{i=1}^{n_j} y_{ij} \text{ mean of group } j$
-   $\overline{y}_{\cdot\cdot} = \frac{1}{n} \sum_{j=1}^a \sum_{i=1}^{n_j} y_{ij} \text{ total mean}$

Solving the problem we have that:

$$\begin{split}
  \sum_{i=1}^{n_j} -2(y_{ij} - \hat{\mu} - \hat{\beta}_j) & = 0 \\
  \sum_{i=1}^{n_j} y_{ij} & = n_j(\hat{\beta}_j + \hat{\mu}) \\
  \overline{y}_{\cdot j} - \hat{\mu} & = \hat{\beta}_j
  \end{split}$$

Also because $\beta_1 = 0$ we have that $\overline{y}_{\cdot 1} - \hat{\mu} = 0$ so $\hat{\mu} = \overline{y}_{\cdot 1}$. Therefore we have that:

-   $\hat{\mu} = \overline{y}_{\cdot 1}$
-   $\hat{\beta}_j = \overline{y}_{\cdot j} - \overline{y}_{\cdot 1} \text{ for } j = 2, \ldots, a$
-   $\hat{\mu}_j = \overline{y}_{\cdot j} \text{ for } j = 1, \ldots, a$

Finally we have that $\hat{\sigma}^2$ is given by:

$$\begin{split}
  \hat{\sigma}^2 & = \frac{1}{n-a} \sum_{j=1}^{a} \sum_{i=1}^{n_j} (y_{ij} - \hat{\mu}_j)^2 \\
  & = \frac{1}{n-a} \sum_{j=1}^{a} (n_j - 1) \sum_{i=1}^{n_j} \frac{(y_{ij} - \hat{\mu}_j)^2}{n_j - 1} \\
  & = \frac{1}{n-a} \sum_{j=1}^{a} (n_j - 1)s_j^2
  \end{split}$$

### Decomposition of variance

-   SST: Sum of Squares Total
-   SSB: Sum of Squares Between
-   SSW: Sum of Squares Within

$$\begin{split}
  SST & = \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot\cdot})^2 \\
  & = \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j} + \overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2 \\
  & = \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2 + 2(y_{ij} - \overline{y}_{\cdot j})(\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot}) + (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2 \\
  & = \sum_{j=1}^a \sum_{i=1}^{n_j} (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2 + \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2 + 2\sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})(\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot}) \\
  \end{split}$$

We have the following result:

$$\begin{split}
  \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})(\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot}) & = \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij}\overline{y}_{\cdot j} - y_{ij}\overline{y}_{\cdot\cdot} - \overline{y}_{\cdot j}^2 + \overline{y}_{\cdot j}\overline{y}_{\cdot\cdot}) \\
  & = \sum_{j=1}^a n_j\overline{y}_{\cdot j}^2 - n\overline{y}_{\cdot\cdot}^2 - \sum_{j=1}^a n_j\overline{y}_{\cdot j}^2 + n_j\overline{y}_{\cdot j}^2 \\
  & = 0
  \end{split}$$

Therefore we have the following result:

$$\begin{split}
  SST & = \sum_{j=1}^a \sum_{i=1}^{n_j} (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2 + \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2 + 2\sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})(\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot}) \\
  & = \sum_{j=1}^a \sum_{i=1}^{n_j} (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2 + \sum_{j=1}^a \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2 \\
  & = SSB + SSW
  \end{split}$$

Now we can define the following hypothesis:

$H_0: \mu_{Room} = \mu_{Cold}$

$H_1: \text{At least one group mean is different from the rest}$

$n_j = [8, 8]$

$\overline{y}_{\cdot j} = [89.375, 176.500]$

$\overline{y}_{\cdot\cdot} = 132.9375$

$F_{1,14} = \frac{\frac{\sum_{j=1}^a \sum_{i=1}^{n_j} (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2}{a-1}}{\frac{\sum_{j=1}^2 \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2}{n-a}} = \frac{\frac{\sum_{j=1}^2 n_j (\overline{y}_{\cdot j} - \overline{y}_{\cdot\cdot})^2}{2-1}}{\frac{\sum_{j=1}^2 \sum_{i=1}^{n_j} (y_{ij} - \overline{y}_{\cdot j})^2}{16-2}} = \frac{\frac{30363.06}{1}}{\frac{1923.875}{14}} = 220.9514$

$Pr[F_{1,14} > 220.9514] = 0.0000000005738688$

$F_{F_{1,14}}^{-1}(p = 0.95) = 4.60011 < 220.9514$

## Two way Anova: aov() & anova()

**Becareful**: Read - <https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-does-the-output-from-anova_0028_0029-depend-on-the-order-of-factors-in-the-model_003f>

-   <https://www.r-bloggers.com/2011/03/anova-%E2%80%93-type-iiiiii-ss-explained/>

-   

We want to explain the variance of the data. First we define the following matrix $\mathbf{M}_{\mathbf{1}_n} \equiv \mathbf{I}_n - \mathbf{1}_n(\mathbf{1}_n^T\mathbf{1}_n)^{-1}\mathbf{1}_n^T$. In general we can define $\mathbf{M}_{X} \equiv \mathbf{I}_n - \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T$. So we have the following result:

$$\begin{split} 
  \mathbf{M}_{\mathbf{1}_n} & = 
  \begin{bmatrix}
  1 & 0 & \cdots & 0 \\ 
  0 & 1 & \cdots & 0 \\ 
  \vdots & \vdots & \ddots & \vdots \\ 
  0 & 0 & \cdots & 1  
  \end{bmatrix}_{n \times n}  - 
  \begin{bmatrix}
  \frac{1}{n} & \frac{1}{n} & \cdots & \frac{1}{n} \\ 
  \frac{1}{n} & \frac{1}{n} & \cdots & \frac{1}{n} \\ 
  \vdots & \vdots & \ddots & \vdots \\ 
  \frac{1}{n} & \frac{1}{n} & \cdots & \frac{1}{n} 
  \end{bmatrix}_{n \times n} \\
  & = \begin{bmatrix}
  1 - \frac{1}{n} & -\frac{1}{n} & \cdots & -\frac{1}{n} \\ 
  -\frac{1}{n} & 1 - \frac{1}{n} & \cdots & -\frac{1}{n} \\ 
  \vdots & \vdots & \ddots & \vdots \\ 
  -\frac{1}{n} & -\frac{1}{n} & \cdots & 1 - \frac{1}{n} 
  \end{bmatrix}_{n \times n}
  \end{split}$$

```{r}
#| echo: true
dim_toy_data <- dim(toy_data)
n_row <- dim_toy_data[1]
one_n_row <- model.matrix(object = TimetoFail ~ 1, data=toy_data)
M_1_n <- (diag(x = n_row) - one_n_row %*% solve(t(one_n_row) %*% one_n_row) 
          %*% t(one_n_row))
M_1_n
```

# References
