---
title: "Identifying Drivers of Outcomes: Linear Models"
author: Luis Francisco Gomez Lopez
institute: FAEDIS
date: 2023-10-31 14:38:43
format:
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    urlcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
editor: visual
bibliography: r_for_marketing_research_and_analytics.bib
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(skimr)
library(tidymodels)
library(dotwhisker)
library(kableExtra)
```

# Contents

-   Please Read Me
-   Purpose
-   Amusement park survey
-   References

# Please Read Me

-   This presentation is based on [@chapman_r_2019, Chapter 7]

# Purpose

-   Apply linear modeling to understand a response variable and make predictions of forecasts

# Amusement park survey

-   **weekend**: whether the visit was on a weekend
-   **num.child**: number of children in the visit
-   **distance**: how far the customer traveled to the park in miles
-   **rides**: satisfaction with rides using a scale \[0, 100\]
-   **games**: satisfaction with games using a scale \[0, 100\]
-   **wait**: satisfaction with waiting times using a scale \[0, 100\]
-   **clean**: satisfaction with cleanliness using a scale \[0, 100\]
-   **overall**: overall satisfaction rating using a scale \[0, 100\]

# Amusement park survey

-   Import data

\tiny

```{r}
#| echo: true
amusement_park <- read_csv("http://goo.gl/HKnl74")
amusement_park |> head(n = 5)
```

# Amusement park survey

-   Transform data

\tiny

```{r}
#| echo: true
amusement_park <- amusement_park |> 
  mutate(weekend = factor(x = weekend, 
                          labels = c('no', 'yes'),
                          ordered = FALSE),
         num.child = as.integer(num.child),
         # logarithmic transform
         logdist = log(distance, base = exp(x = 1)))
amusement_park |> head(n = 5)
```

# Amusement park survey

-   Summarize data

    -   Ups the table is really big!!! Try it in your console to see the complete table

\tiny

```{r}
#| echo: true
amusement_park |> skim()
```

# Amusement park survey

-   Correlation matrices

    -   Pearson correlation coefficients for samples in a tibble

\tiny

```{r}
#| echo: true
#| out-width: 80%
correlation_matrix <- amusement_park |> 
  select(num.child, rides:logdist) |>
  corrr::correlate()
correlation_matrix
```

# Amusement park survey

-   Correlation matrices

    -   Pearson correlation coefficients for samples in a tibble

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
correlation_matrix |> autoplot(triangular = "lower")
```

# Amusement park survey

-   **Bivariate Association: the base R way**

\tiny

```{r}
#| echo: true
#| out-width: 60%
#| fig-align: center
plot(overall~rides, data=amusement_park,
     xlab="Satisfaction with Rides", ylab="Overall Satisfaction")
abline(reg = lm(formula = overall~rides, data = amusement_park), 
       col = 'blue')
```

# Amusement park survey

-   **Bivariate Association: the tidyverse way**

\tiny

```{r}
#| echo: true
#| out-width: 55%
#| fig-align: center
amusement_park |> ggplot(aes(x = rides, y = overall)) +
  geom_point() + 
  geom_smooth(method = 'lm', 
              color = 'blue', 
              se = FALSE) + 
  labs(x = "Satisfaction with Rides",
       y = "Overall Satisfaction")
```

# Amusement park survey

-   Linear Model with a Single Predictor

```{r}
#| echo: false
#| out-width: 65%
#| fig-align: center
# A tutorial related to set.seed
## https://r-coder.com/set-seed-r/
set.seed(12345)
toy_data <- amusement_park |> 
  slice_sample(n = 15)

toy_model <- lm(formula = overall ~ rides, data = toy_data)

augment(x = toy_model) |> 
  ggplot(aes(x = rides, y = overall)) + 
  geom_point(shape = 21, 
             color = 'black', fill = '#E31A1C',
             size = 2) +
  geom_smooth(method = 'lm', 
              color = '#2C3E50',
              se = FALSE) + 
  geom_segment(aes(xend=rides, yend=.fitted),
               color='#CCBE93') + 
  labs(x = "Satisfaction with Rides",
       y = "Overall Satisfaction")
```

# Amusement park survey

-   Linear Model with a Single Predictor

$$overall_{i} = \beta_0 + \beta_1 rides_i + \epsilon_i \text{ where } \epsilon_i \sim \mathcal{N}(0, \sigma^2) \text{ and } i = 1, \ldots, 500$$ $$\widehat{overall}_{i} = \widehat{\beta}_0 + \widehat{\beta}_1 rides_i \text{ and } \widehat{\sigma}^2 \text{ where } i = 1, \ldots, 500$$ $$overall_{i} - \widehat{overall}_{i} = \widehat{\epsilon}_i \text{ where } i = 1, \ldots, 500$$

\tiny

```{r}
#| echo: true
model1 <- lm(formula = overall ~ rides, data = amusement_park)
model1
```

# Amusement park survey

-   Linear Model with a Single Predictor

\tiny

```{r}
#| echo: true
ls.str(model1)
```

# Amusement park survey

-   Linear Model with a Single Predictor

\tiny

```{r}
#| echo: true
summary(model1)
```

# Amusement park survey

-   Linear Model with a Single Predictor

\tiny

```{r}
#| echo: true
model1$coefficients
# Make some predictions
# We want to forecast the overall satisfaction rating
# if the satisfaction with rides is 95
-94.962246 + 1.703285*95
```

# Amusement park survey

-   Linear Model with a Single Predictor

    -   Std. Error column

        -   Indicates uncertainty in the coefficient estimate
        -   We can build a confidence interval

\tiny

```{r}
#| echo: true
summary(model1)$coefficients[, 2]
confint(model1, level = 0.95)
```

# Amusement park survey

-   Linear Model with a Single Predictor

\tiny

```{r}
#| echo: true
#| out-width: 60%
#| fig-align: center
par(mfrow=c(2,2))
plot(model1)
par(mfrow=c(1,1))
```

# Amusement park survey

-   Linear Model with a Single Predictor

    -   **Linearity**: plot $(1,1)$

        -   Reference line should be flat and horizontal

    -   **Normality of residuals**: plot $(1,2)$

        -   Dots should fall along the line

    -   **Homogeneity of variance**: plot $(2,1)$

        -   Reference line should be flat and horizontal

    -   **Influential observations**: plot $(2,2)$

        -   Points should be inside the contour lines

# Amusement park survey

-   Linear Model with Multiple Predictors

$$\begin{split}
   overall_{i} & = \beta_0 + \beta_1 rides_i + \beta_2 games_i \\
   & + \beta_3 wait_i + \beta_4 clean_i + \epsilon_i \\
   & \text{ where } \epsilon_i \sim \mathcal{N}(0, \sigma^2) \text{ and } i = 1, \ldots, 500
   \end{split}$$

\tiny

```{r}
#| echo: true
model2 <- lm(formula = overall ~ rides + games + wait + clean, 
             data = amusement_park)
model2
```

# Amusement park survey

-   Linear Model with Multiple Predictors

\tiny

```{r}
#| echo: true
#| out-width: 60%
#| fig-align: center
par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))
```

# Amusement park survey

-   Linear Model with Multiple Predictors

\tiny

```{r}
#| echo: true
summary(model2)
```

# Amusement park survey

-   Linear Model with Multiple Predictors

$H_0: \beta_1 = 0$

$H_1: \beta_1 \neq 0$

$t_{rides} = \frac{\hat{\beta}_1 - \beta_1}{\sqrt{\widehat{Var(\hat{\beta}_1)}}} = \frac{0.529078 - 0}{0.14207176} = 3.724019$

\tiny

```{r}
#| echo: true
model2$coefficients
# Calculate the variance-covariance matrix, extract
# the diagonal and calculate the standard deviaton of
# the parameters
model2 |> vcov() |> diag() |> sqrt()
```

# Amusement park survey

-   Linear Model with Multiple Predictors

```{r}
#| echo: false
#| out-width: 70%
#| fig-align: center
t_test_rides <- summary(model2)$coefficients[, 3][2]
degrees_of_freedom <- summary(model2)$df[2]

ggplot() + 
  geom_function(fun=dt, args=list(df=degrees_of_freedom),
                xlim=c(-5,5),
                color='#2C3E50') +
  geom_vline(xintercept = qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                             lower.tail = TRUE),
             color="#E31A1C") +
  geom_vline(xintercept = qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                             lower.tail = FALSE),
             color="#E31A1C") +
  geom_vline(xintercept = t_test_rides |> round(digits = 2) |> unname(),
             color="#18BC9C") +
  geom_ribbon(data = tibble(x = seq.int(from = -5,
                                        to = qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                                lower.tail = TRUE),
                                        by = 0.01),
                            y = dt(x = x, df = degrees_of_freedom, ncp = 0)),
              aes(x=x, ymin = 0, ymax=y),
              fill='#CCBE93', alpha=0.5) +
  geom_ribbon(data = tibble(x = seq.int(from = qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                                  lower.tail = FALSE),
                                        to = 5,
                                        by = 0.01),
                            y = dt(x = x, df = degrees_of_freedom, ncp = 0)),
              aes(x=x, ymin = 0, ymax=y),
              fill='#CCBE93', alpha=0.5) +
  scale_x_continuous(breaks = c(-5,
                                t_test_rides,
                                qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                   lower.tail = TRUE),
                                qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                   lower.tail = FALSE),
                                5),
                     labels = scales::label_number(accuracy = 0.01)) +
  labs(x=NULL,
       y=NULL,
       color=NULL,
       title = "Student's t-distribution distribution function",
       subtitle = str_glue('Degrees of freedom= {summary(model2)$df[2]},
                           Critical values: ({qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                                lower.tail = TRUE) |> round(digits=2)},{qt(p = 0.025, df = degrees_of_freedom, ncp = 0, 
                                                lower.tail = FALSE) |> round(digits=2)})
                           T statistic: {t_test_rides |> round(digits=2)}')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.position   = 'bottom', 
        axis.text         = element_text(face = "bold"))
```

# Amusement park survey

-   Linear Model with Multiple Predictors

\tiny

```{r}
#| echo: true
confint(model2, level = 0.95)
```

# Amusement park survey

-   Linear Model with Multiple Predictors

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
library(coefplot) # Remember to install the package if it is not installed
coefplot(model = model2, 
         # The intercept is relatively large: -131.4092 
         intercept = FALSE,
         ylab="Rating of Feature", 
         xlab="Association with Overall Satisfaction",
         lwdOuter = 1.5)
```

# Amusement park survey

-   Comparing models

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
summary(model1)$r.squared
summary(model2)$r.squared
summary(model1)$adj.r.squared
summary(model2)$adj.r.squared
```

# Amusement park survey

-   Comparing models

    -   **Base R way**

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
plot(x = amusement_park$overall, y = fitted(model1),
     col = "red", xlim = c(0,100), ylim = c(0,100),
     xlab = "Actual Overall Satisfaction", 
     ylab = "Fitted Overall Satisfaction")
points(x = amusement_park$overall, y = fitted(model2),
       col = "blue")
legend(x = "topleft", legend = c("Model 1", "Model 2"), col = c("red", "blue"), pch = 1)
```

# Amusement park survey

-   Comparing models

    -   **Tidymodels and tidyverse way**: Prepare data

\tiny

```{r}
#| echo: true
model1_augment <- augment(x = model1) |> mutate(model = "Model 1")
model2_augment <- augment(x = model2) |> mutate(model = "Model 2")
models_performance <- model1_augment |> bind_rows(model2_augment)

models_performance |> glimpse()
```

# Amusement park survey

-   Comparing models

    -   **Tidymodels and tidyverse way**: Visualize

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
models_performance |> 
  ggplot() +
  geom_point(aes(x = overall, y = .fitted,
                 color = model)) +
  labs(x = "Actual Overall Satisfaction",
       y = "Fitted Overall Satisfaction")
```

# Amusement park survey

-   Comparing models

    -   Analysis of variance (`anova`) for nested models[^1]

[^1]: This statistical analysis only make sense for nested models that are fitted with the same data where the convention is to include the models from smallest to largest. See `?anova.lm`

\tiny

```{r}
#| echo: true
anova_lm <- anova(model1, model2, test = "F")
anova_lm
```

# Amusement park survey

-   Comparing models

\tiny

$H_0: \beta_0 = \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$

$H_1: \text{At least one } \beta_j \neq 0 \text{ for } j = 0, 1, 2, 3, 4$

$F = \frac{\frac{RSS_1 - RSS_2}{p_2 - p_1}}{\frac{RSS_2}{n - p_2}} = \frac{\frac{82611.81 - 55531.53}{5 - 2}}{\frac{55531.53}{500 - 5}} = 80.46323$

```{r}
#| echo: false
#| out-width: 55%
#| fig-align: center

df1 <- 3
df2 <- 495
f_statistic <- anova_lm |> 
  tidy() |>
  select(statistic) |> 
  pull(1) |> 
  _[2]

ggplot() + 
  geom_function(fun=df, args=list(df1=df1, df2=df2, log=FALSE),
                xlim=c(0,90),
                color='#2C3E50') +
  geom_ribbon(data = tibble(x = seq.int(from = qf(p = 0.05,
                                                  df1 = df1,
                                                  df2 = df2,
                                                  lower.tail = FALSE),
                                        to = 90,
                                        by = 0.01),
                            y = df(x = x, df1 = df1, df2 = df2)),
              aes(x = x, ymin = 0, ymax = y),
              fill='#E31A1C',
              alpha=0.1) +
  geom_vline(xintercept = qf(p = 0.05, df1=df1, df2=df2, lower.tail = FALSE),
             color="#E31A1C") +
  geom_vline(xintercept = f_statistic,
             color="#18BC9C") +
               scale_x_continuous(breaks = c(qf(p = 0.05, df1 = df1, df2=df2, lower.tail = FALSE),
                                             f_statistic,
                                             90),
                                             labels = scales::label_number(accuracy = 0.1)) +
                                    labs(x=NULL,
                                         y=NULL,
                                         color=NULL,
                                         title = 'F-squared distribution function',
                                         subtitle = str_glue('df1={df1},
                           df2={df2},
                           Critical value: {qf(p = 0.05, df1=3, df2=495, lower.tail = FALSE) |> round(digits=2)}
                           F statistic: {f_statistic |> round(digits=2)}')) +
                                    theme(panel.border      = element_rect(fill = NA, color = "black"),
                                          plot.background   = element_rect(fill = "#f3fcfc"),
                                          panel.background  = element_rect(fill = "#f3f7fc"),
                                          plot.title        = element_text(face = "bold"),
                                          axis.title        = element_text(face = "bold"),
                                          legend.title      = element_text(face = "bold"),
                                          legend.position   = 'bottom',
                                          axis.text         = element_text(face = "bold"))
```

# Amusement park survey

-   Predictions

$$\begin{split}
   \widehat{overall}_{j} & = \widehat{\beta}_0 + \widehat{\beta}_1 rides_j + \widehat{\beta}_2 games_j \\
   & + \widehat{\beta}_3 wait_j + \widehat{\beta}_4 clean_j
   \end{split}$$

\tiny

```{r}
#| echo: true
coef(model2) |> enframe(name = "coef")
```

# Amusement park survey

-   Predictions

    -   Manual

\tiny

```{r}
#| echo: true
(coef(model2)["(Intercept)"]*1 + coef(model2)["rides"]*30 + coef(model2)["games"]*10 + 
    coef(model2)["wait"]*57 +  coef(model2)["clean"]*90) |>
  unname()
```

-   Predictions

    -   Matrix multiplication

```{r}
#| echo: true
coef(model2) %*% c(1, 30, 10, 57, 90)
```

# Amusement park survey

-   Predictions

    -   `predict`

\tiny

```{r}
#| echo: true
# New data
new_data <- tibble(rides = c(30, 70),
                   games = c(10, 80),
                   wait =  c(57, 60),
                   clean = c(90, 93))
# Result
predict(object = model2, newdata = new_data) |> 
  enframe(name = "observation", value = "overall_pred") |>
  bind_cols(new_data)
```

# Amusement park survey

-   Standardizing the predictors

    -   Compare the effect that different predictor variables have on a response variable
    -   It must be interpreted in terms of standard deviations
        -   One standard deviation in $x$ variable is associated with a standard deviation increase of decrease depending on the value of the estimated parameter

\tiny

```{r}
#| echo: true
amusement_park_std <- amusement_park |> 
  select(-distance) |> 
  mutate(across(rides:logdist, 
                .fns = ~ scale(x = .x, 
                               center = TRUE, 
                               scale = TRUE)[,1]))
amusement_park_std |> head()
```

# Amusement park survey

-   Standardizing the predictors

\tiny

```{r}
#| echo: true
model2_std <- lm(formula = overall ~ rides + games + wait + clean, 
             data = amusement_park_std)
summary(model2_std)
```

# Amusement park survey

-   Using factors as predictors

\tiny

```{r}
#| echo: true
model3 <- lm(formula = overall ~ rides + games + wait + clean + weekend + logdist + num.child,
             data = amusement_park_std)
tidy(model3)
```

```{r}
#| echo: true
glance(model3)
```

# Amusement park survey

-   Using factors as predictors

    -   Overall satisfaction is about the same regardless the number of children

\tiny

```{r}
#| echo: true
amusement_park_std <- amusement_park_std |> 
  mutate(num.child.factor = factor(num.child))
model4 <- lm(formula = overall ~ rides + games + wait + clean + weekend + logdist + num.child.factor,
             data = amusement_park_std)
tidy(model4) |> slice(1, 2, 8:12)
```

```{r}
#| echo: true
glance(model4)
```

# Amusement park survey

-   Using factors as predictors

    -   Preparing data

\tiny

```{r}
#| echo: true
amusement_park_std <- amusement_park_std |>
  mutate(has.child = factor(x = num.child > 0, labels = c("No", "Yes")))
model5 <- lm(formula = overall ~ rides + games + wait + clean + logdist + has.child,
             data = amusement_park_std)
tidy(model5) |> slice(1, 2, 7)
```

```{r}
#| echo: true
glance(model5)
```

# Amusement park survey

-   Using factors as predictors

    -   Maybe having children and the visits on weekends are important for the scores so an interaction will be useful

\tiny

```{r}
#| echo: true
model6 <- lm(formula = overall ~ rides + games + wait + clean + weekend + logdist + 
                                 has.child + rides:has.child + games:has.child + wait:has.child + 
                                 clean:has.child + rides:weekend + games:weekend + wait:weekend + 
                                 clean:weekend, data = amusement_park_std)
tidy(model6) |> slice(9:16)
glance(model6)
```

# Amusement park survey

-   Using factors as predictors

    -   Only an interaction was significant

\tiny

```{r}
#| echo: true
model7 <- lm(formula = overall ~ rides + games + wait + clean + logdist + has.child +
                       wait:has.child, data = amusement_park_std)
tidy(model7)
glance(model7)
```

# Amusement park survey

-   Using factors as predictors

    -   Final model

\tiny

```{r}
#| echo: true
#| out-width: 50%
#| fig-align: center
library(dotwhisker)  # Remember to install the package if it is not installed
tidy(model7) |> 
  dwplot(ci = 0.95,
         dot_args = list(size = 2, color = "black"), whisker_args = list(color = "red"),
         vline = geom_vline(xintercept = 0, color = "black", linetype = 2)) + 
  labs(x = "Association with Overall Satisfaction", y = "Rating of Feature")
```

# Amusement park survey

-   Formula syntax

```{r}
tibble(formula = c("y $\\sim$ x",
                   "y $\\sim$ -1 + x",
                   "y $\\sim$ x + z",
                   "y $\\sim$ x + z + x:z",
                   "y $\\sim$ x*z",
                   "y $\\sim$ (x + z + w)$\\^{}$2",
                   "y $\\sim$ (x + z + w)$\\^{}$2 - x:z",
                   "y $\\sim$ x + I(x$^2$)"),
       model = c("$y_i = \\beta_0 + \\beta_1x_i + \\varepsilon_i$",
                 "$y_i = \\beta_1 x_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1 x_i + \\beta_2 z_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1 x_i + \\beta_2 z_i + \\beta_3 x_i z_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1 x_i + \\beta_2 z_i + \\beta_3 x_i z_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1 x_i + \\beta_2 z_i + \\beta_3 w_i + \\beta_4 x_i z_i + \\beta_5 x_i w_i + \\beta_6 w_i z_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1 x_i + \\beta_2 z_i + \\beta_3 w_i + \\beta_4 x_i w_i + \\beta_5 w_i z_i + \\varepsilon_i$",
                 "$y_i = \\beta_0 + \\beta_1x_i + \\beta_1x_i^2 + \\varepsilon_i$")) |> 
  kbl(format = "latex", 
      escape = FALSE, booktabs = TRUE, 
      col.names = c("Formula in R", "Statistical Model"),
      linesep = "") |> 
  kable_styling(position = "center", font_size = 7)
```

-   Try the following models using `tidy`:

\tiny

```{r}
#| echo: true
#| eval: false
lm(formula = overall ~ rides, data = amusement_park_std) |> tidy()
lm(formula = overall ~ -1 + rides, data = amusement_park_std) |> tidy()
lm(formula = overall ~ rides + has.child, data = amusement_park_std) |> tidy()
lm(formula = overall ~ rides + has.child + has.child, data = amusement_park_std) |> tidy()
lm(formula = overall ~ (rides + has.child + weekend)^2, 
   data = amusement_park_std) |> tidy()
lm(formula = overall ~ (rides + has.child + weekend)^2 - rides:has.child, 
   data = amusement_park_std) |> tidy()
lm(formula = overall ~ rides + I(rides^2) - rides:has.child, data = amusement_park_std) |> tidy()
```

# References
