---
title: "Segmentation: Clustering"
author: Luis Francisco Gomez Lopez
institute: FAEDIS
date: 2024-03-16 23:06:47
format:
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    linkcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
editor: visual
bibliography: r_for_marketing_research_and_analytics.bib
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(skimr)
```

# Contents

-   Please Read Me
-   Purpose
-   Consumer segmentation survey
-   
-   References

# Please Read Me

-   This presentation is based on [@chapman_r_2019, Chapter 11]

# Purpose

-   

# Consumer segmentation survey

-   **age**: age of the consumer in years
-   **gender**: if the consumer is male of female
-   **income**: yearly disposable income of the consumer
-   **kids**: number of children of the consumer
-   **ownHome**: if the consumer owns a home
-   **subscribe**: if the consumer is subscribed or not

# Consumer segmentation survey

-   **Import data**

\tiny

```{r}
#| echo: true
segmentation <- read_csv(file = "http://goo.gl/qw303p") |> 
  select(-Segment)
segmentation |> head(n = 5)
```

# Consumer segmentation survey

-   **Inspect data**

\tiny

```{r}
#| echo: true
segmentation |> glimpse()
```

# Consumer segmentation survey

-   **Summarize data**

    -   Ups the table is really big!!! Try it in your console to see the complete table

\tiny

```{r}
#| echo: true
segmentation |> skim()
```

# References

