---
title: "Specific instructions: marketing report part 1"
execute: 
  echo: true
  warning: false
format: 
  html:
    number-sections: true
    toc: true
    embed-resources: true
    css: ../000_css/styles.css
bibliography: ../000_references/marketing_research_and_analytics.bib
editor: source
---

# Business Case

[Disclaimer:]{style="color: red;"} This business case was taken from the course [Data Science for Business Part 1](https://university.business-science.io/p/ds4b-101-r-business-analysis-r) offered by the company [Business Science](https://www.business-science.io/) and adapted to be in line with the topics covered in [@chapman_r_2019].

You and your team work for a corporation located in Wilton, Connecticut, United States that supplies bicycle frames and other components related to bicycles to different bicycle shops through the United States.

Your team is assigned to complete 2 tasks:

-   Support the Research and Development (R & D) division in identifying potential new products and pricing them by using data collected from the bicycle shops.

-   Support the marketing team in the creation of a marketing segmentation clustering model by using data collected from to the bicycle shops to offer more personalized products and messaging them.

# Data

The data collected by the corporation is available in the `sweep` package

-   Load the library: `library(sweep)`
-   Type `bike_sales` to access the data
-   If you want information about this data set type `?bike_sales`

The `bike_sales` data set contains 17 variables and a total of 15644 observations relate to transactional data:

```{r}
library(sweep)
library(tidyverse)
library(DT)

dim(bike_sales)

bike_sales |>
  head() |>
  datatable()
```

Each observation corresponds to a closed order. A closed order is a completed transaction where all items have been shipped and invoiced to a bicycle shop. A closed order, represented in the `order.id`, can be associated to 1 or more transactions and are capture by the `order.line` variable.

For example, the first 2 rows corresponds to a closed order where a bike shop named **Ithaca Mountain Climbers** bought 2 models: **Jekyll Carbon 2** and **Trigger Carbon 2**. The `order.id` is identify as `1` and the `order.line` variable has a value of `1` and `2` taking into account that `1` (see the `quantity` variable) **Jekyll Carbon 2** and `1` (see the `quantity` variable) **Trigger Carbon 2** model were acquired at a price of `6070` and `5970`.

```{r}
bike_sales |>
  select(
    order.date,
    order.id,
    order.line,
    quantity,
    price,
    customer.id,
    bikeshop.name,
    model
  ) |>
  datatable()
```

You can use the following command inside RStudio IDE to inspect all the columns as a spreadsheet:

```{r}
#| eval: false
bike_sales |>
  View()
```

# Marketing report structure part 1

In this first report you are going to describe the `bike_sales` data as a first step to complete the 2 tasks.

Use the template `001_marketing_report_part-1.qmd` to build the marketing report. You can find the template to use in 2 locations. Choose the option of your preference:

-   <https://github.com/luifrancgom/marketing_research_and_analytics/> \> 000_templates \> 001_marketing_report_part-1.qmd

-   Primer Corte 30% \> Learning Resources \> Support material \> Templates \> 001_marketing_report_part-1.qmd

# References