---
title: An Overview of the R Language
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: ../000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: true
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: ../000_references/marketing_research_and_analytics.bib
csl: ../000_references/apa-7.csl
editor: source
---

```{r}
#| label: libraries
#| echo: false

```

```{r}
#| label: palette
#| echo: false

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   This presentation is based on [@chapman_r_2019, Chapter 2]

## 

- **Purpose**

    - Equip beginners with a programming foundation by leveraging the R language, enabling practical application in marketing research and analytics.
    
# Sales and product satisfaction survey

##

-   **Ordinal 7 point scale**

    -   Extremely satisfied: 7
    -   Moderately satisfied: 6
    -   Slightly satisfied: 5
    -   Neither satisfied or dissatisfied: 4
    -   Slightly dissatisfied: 3
    -   Moderately dissatisfied: 2
    -   Extremely dissatisfied: 1

##

-   **Variables**

    -   iProdSAT: satisfaction with a product
    -   iSalesSAT: satisfaction with sales experience
    -   iProdREC: likelihood to recommend the product
    -   iSalesREC: likelihood to recommend the sales person
    -   Segment: market segment assigned by a clustering algorithm [@chapman_r_2019, Chapter 11]

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|3-5|"
library(tidyverse) # Remember to load the tidyverse library

satisfaction_data <- read_csv(
  file = "http://goo.gl/UDv12g"
)

satisfaction_data |>
  head(n = 5)
```
:::
::: {.column width="40%"}
-  **Import data**

   - The tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|4-9|"
library(skimr)

satisfaction_data <- satisfaction_data |>
  mutate(
    Segment = factor(
      x = Segment,
      ordered = FALSE
    )
  )
```
\
```{r}
#| eval: false
#| code-line-numbers: "|2|"
satisfaction_data |>
  skim()
```
:::
::: {.column width="50%"}
-  **Transform data**

   - The tidyverse way
\
\
\
\
\
-  **Using `skimr`**

   - Ups the table is really big!!! Try it in your console to see the complete table
::: 
::::

##

-   **R objects**: everything in R is an object (object-oriented). For now, we will only inspect a few selected objects:

    -   **Atomic vectors**[^1]

        -   Logical
        -   Integer
        -   Double

    -   **Factors**

    -   **Data Frames**

    -   **Tibbles**

[^1]: In R the atomic vectors are logical, integer, double, numeric (which includes integer and double), character, complex and raw but for pedagogical purposes we are going to check later character, numeric includes integer and double and we are not going to use complex and raw

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: true
as.integer(satisfaction_data$Segment)[1:5] == 1
```
\
```{r}
#| echo: true
as.integer(satisfaction_data$Segment)[1:5] > 1
```
\
```{r}
#| echo: true
as.integer(satisfaction_data$Segment)[1:5] >= 1
```
:::
::: {.column width="50%"}
-  **Atomic vectors**
   -  Logical
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
```{r}
as.integer(satisfaction_data$Segment)[1:5]
```
\
\
\
```{r}
sprintf("%.2f", satisfaction_data$iProdSAT[1:5])
```
:::
::: {.column width="45%"}
-   **Atomic vectors**

    -   Integer
\
\
\
-   **Atomic vectors**

    -   Double
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
satisfaction_data$Segment[1:5]
```
\
\
```{r}
#| echo: true
as.data.frame(satisfaction_data) |>
  head(n = 5)
```
:::
::: {.column width="50%"}
-   **Factors**
\
\
\
\
-   **Data Frames**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
satisfaction_data |>
  head(n = 5)
```
\
\
```{r}
class(satisfaction_data)
```
:::
::: {.column width="50%"}
-  **Tibble**
   - The tidyverse way
\
\
\
\
\
\
\
-  Checking the object class of a **tibble**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
```{r}
#| code-line-numbers: "|2-4|"
satisfaction_data <- satisfaction_data |>
  mutate(
    customer = 1:nrow(satisfaction_data)
  ) |>
  head(n = 5)

satisfaction_data
```
:::
::: {.column width="45%"}
-  **Add new variables**

   - The tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
```{r}
#| code-line-numbers: "|2-9|"
satisfaction_data |>
  select(
    customer,
    Segment,
    iProdSAT,
    iSalesSAT,
    iProdREC,
    iSalesREC
  ) |>
  head(n = 5)
```
:::
::: {.column width="45%"}
-  **Picks variables based on their names**

   - The tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
```{r}
#| code-line-numbers: "|2-4|"
satisfaction_data |>
  filter(
    Segment == 2
  ) |>
  head(n = 5)
```
:::
::: {.column width="45%"}
-  **Picks cases based on their values**

   - The tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|2-5|"
satisfaction_data |>
  summarise(
    mean_iProdSAT = mean(iProdSAT),
    median_iSalesSAT = median(iSalesSAT)
  )
```
:::
::: {.column width="50%"}
-  **Reduces multiple values to a single summary**

   - The tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|2|3|4-7|"
satisfaction_data |>
  group_by(Segment) |>
  select(iProdSAT, iSalesSAT) |>
  summarise(
    mean_iProdSAT = mean(iProdSAT),
    mean_iSalesSAT = mean(iSalesSAT)
  )
```
:::
::: {.column width="50%"}
-  **Does product and sales satisfaction differ by segment?**

   - The tidyverse way
:::
::::

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/products/open-source/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**skimr**](https://CRAN.R-project.org/package=skimr) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}