---
title: Relationships Between Continuous Variables
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: ../000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: true
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: ../000_references/marketing_research_and_analytics.bib
csl: ../000_references/apa-7.csl
editor: source
---

```{r}
#| label: libraries
#| echo: false
library(tidyverse)
library(skimr)
```

```{r}
#| label: palette
#| echo: false
umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   This presentation is based on [@chapman_r_2019, Chapter 4]

## 

- **Purpose**

    - Understand the relationships between pairs of variables in multivariate data and examine how to visualize the relationships and compute statistics that describe their associations

# CRM system data

## {.smaller}

-   **cust.id**: customer identifier
-   **age**: decimal age in years
-   **credit.score**: 3-digit number in \[300, 900\], representing the credit risk
-   **email**: whether or not there is information about the customer email
-   **distance.to.store**: distance in kilometers to the nearest physical store
-   **online.visits**: yearly visits to the online store
-   **online.trans**: yearly online orders
-   **online.spend**: yearly spending in those online orders
-   **store.trans**: yearly orders in physical stores
-   **store.spend**: yearly spending in those physical store orders

## {.smaller}

-   **sat.service**: satisfaction with service using an ordinal 5 point scale and collected using a survey

-   **sat.selection**: satisfaction with product selection using an ordinal 5 point scale and collected using a survey

    -   **Ordinal 5 point scale used and possible values in the survey**:

        -   Extremely satisfied: 5
        -   Very satisfied: 4
        -   Moderately satisfied: 3
        -   Very unsatisfied: 2
        -   Extremely unsatisfied: 1
        -   NA: customer did not response the survey

## {.smaller .center}

:::: {.columns}
::: {.column width="75%"}
```{r}
#| code-line-numbers: "|1-3|"
customer <- read_csv(
  file = "http://goo.gl/PmPkaG"
)

customer |> head(n = 5)
```
:::
::: {.column width="25%"}
-   **Import data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="75%"}
```{r}
#| code-line-numbers: "|2-10|"
customer <- customer |>
  mutate(
    cust.id = factor(x = cust.id, ordered = FALSE),
    email = factor(x = email, ordered = FALSE),
    online.visits = as.integer(x = online.visits),
    online.trans = as.integer(x = online.trans),
    store.trans = as.integer(x = store.trans),
    sat.service = factor(x = sat.service, ordered = TRUE),
    sat.selection = factor(x = sat.selection, ordered = TRUE)
  )

customer |> head(n = 5)
```
:::
::: {.column width="25%"}
-   **Transform data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="70%"}
```{r}
#| code-line-numbers: "|2|"
customer |>
  glimpse()
```
:::
::: {.column width="30%"}
-   **Inspect data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
#| code-line-numbers: "|2|"
customer |>
  skim()
```
:::
::: {.column width="50%"}
-   **Summarize data**

    -   Ups the table is really big!!! Try it in your console to see the complete table
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
- **Scatterplots**

  - the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|3-9|"
customer |>
  ggplot() +
  geom_point(
    aes(
      x = store.spend,
      y = online.spend,
      color = email
    )
  )
```
:::
::: {.column width="50%"}
\
\
```{r}
#| label: fig-geom-point
#| fig-cap: Using `geom_point()` to apply a geometry
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(
    aes(
      x = store.spend,
      y = online.spend,
      color = email
    )
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
- **Scatterplots**

  - the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|8|11-17|18-24|"
customer |>
  ggplot() +
  geom_point(
    aes(
      x = store.spend,
      y = online.spend,
      color = email,
      shape = email
    )
  ) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c(
      "email on file: no",
      "email on file: yes"
    )
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c(
      "email on file: no",
      "email on file: yes"
    )
  )
```
:::
::: {.column width="45%"}
\
\
```{r}
#| label: fig-scale-color-shape-manual
#| fig-cap: Using `scale_color_manual()` and `scale_shape_manual()` to modify scales for the `color` and `shape` aesthetics
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(
    aes(
      x = store.spend,
      y = online.spend,
      color = email,
      shape = email
    )
  ) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c(
      "email on file: no",
      "email on file: yes"
    )
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c(
      "email on file: no",
      "email on file: yes"
    )
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
- **Scatterplots**

  - the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|17-20|21-24|"
customer |>
  ggplot() +
  geom_point(aes(
    x = store.spend,
    y = online.spend,
    color = email,
    shape = email
  )) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(1, 2, 5, 10, 20, 50, 100, 500)
  ) +
  scale_y_continuous(
    transform = "log1p",
    breaks = c(1, 5, 50, 500)
  )
```
:::
::: {.column width="45%"}
\
\
```{r}
#| label: fig-scale-continuous
#| fig-cap: Using `scale_x_continuous()` and `scale_y_continuous()` to modify scales for the `x` and `y` coordinates
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(aes(
    x = store.spend,
    y = online.spend,
    color = email,
    shape = email
  )) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(1, 2, 5, 10, 20, 50, 100, 500)
  ) +
  scale_y_continuous(
    transform = "log1p",
    breaks = c(1, 5, 50, 500)
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
- **Scatterplots**

  - the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|22-26|"
customer |>
  ggplot() +
  geom_point(aes(
    x = store.spend,
    y = online.spend,
    color = email,
    shape = email
  )) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_x_continuous(
    transform = "log1p",
    breaks = c(1, 2, 5, 10, 20, 50, 100, 500)
  ) +
  scale_y_continuous(trans = "log1p", breaks = c(1, 5, 50, 500)) +
  labs(
    x = "Prior 12 months in-store sales ($)",
    y = "Prior 12 months online sales ($)",
    title = "Customers as of June 2014"
  )
```
:::
::: {.column width="40%"}
\
\
```{r}
#| label: fig-labs-scatter
#| fig-cap: Using `labs()` to modify the labels
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(aes(
    x = store.spend,
    y = online.spend,
    color = email,
    shape = email
  )) +
  scale_color_manual(
    values = c("black", "green3"),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_shape_manual(
    values = c(1, 19),
    labels = c("email on file: no", "email on file: yes")
  ) +
  scale_x_continuous(
    transform = "log1p",
    breaks = c(1, 2, 5, 10, 20, 50, 100, 500)
  ) +
  scale_y_continuous(trans = "log1p", breaks = c(1, 5, 50, 500)) +
  labs(
    x = "Prior 12 months in-store sales ($)",
    y = "Prior 12 months online sales ($)",
    title = "Customers as of June 2014"
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="70%"}
$$
r_{xy} = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2}\sqrt{\sum_{i=1}^n (y_i - \bar{y})^2}}
$$ {#eq-pearson-cor-coef}

- $n$ is the sample size
- we must have paired numeric data $\{ (x_1, y_1), ..., (x_n, y_n) \}$
- $\bar{x} = \sum_{i=1}^n x_i$
- $\bar{y} = \sum_{i=1}^n y_i$
:::
::: {.column width="30%"}
-   **Correlation Coefficients**

    -   Pearson correlation coefficient for a sample
    -   @eq-pearson-cor-coef is a "nasty" formula but we can brake it down in smaller chunks
:::
::::

## {.smaller .center}

:::: {.columns}

::: {.column width="70%"}
```{r}
age_mean <- mean(customer$age)
age_credit.score <- mean(customer$credit.score)
numerator <- sum(
  (customer$age - age_mean) * (customer$credit.score - age_credit.score)
)
denominator <- sqrt(sum((customer$age - age_mean)^2)) *
  sqrt(sum(((customer$credit.score - age_credit.score)^2)))
pearson_corr <- numerator / denominator
pearson_corr
```
\
\
```{r}
#| classes: fragment
cor(
  customer$age,
  customer$credit.score,
  method = "pearson"
)
```
:::

::: {.column width="30%"}
-   **Correlation Coefficients**

    -   Pearson correlation coefficient for a sample

\
\
\

::: {.fragment}
-   **But don't worry be happy!!!**

    -   Use `cor`
:::
:::

::::

## {.smaller .center}

:::: {.columns}
::: {.column width="80%"}
```{r}
#| code-line-numbers: "|5-9|"
library(corrr) # Remember to install the package if it is not installed

correlation_matrix <- customer |>
  select(where(is.numeric)) |>
  correlate(
    use = "pairwise.complete.obs", # There are NA values
    method = "pearson",
    diagonal = NA
  )

correlation_matrix # Ups!!! The tibble is wide. Check out the tibble in your console
```
:::
::: {.column width="20%"}
-   **Correlation matrices**

    -   Pearson correlation coefficients for samples in a tibble
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   **Correlation matrices**

    -   Pearson correlation coefficients for samples in a tibble
```{r}
#| eval: false
#| code-line-numbers: "|2|"
correlation_matrix |>
  autoplot(triangular = "lower")
```
:::
::: {.column width="50%"}
\
\
\
```{r}
#| label: fig-correlation-matrix-plot
#| fig-cap: Visualizing a correlation matrix using `corrr`
#| echo: false
#| lightbox: true
correlation_matrix |>
  autoplot(triangular = "lower")
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Transforming variables**

```{r}
cor(customer$store.spend, customer$distance.to.store)
```
\
```{r}
#| eval: false
customer |>
  ggplot() +
  geom_point(
    aes(
      x = distance.to.store,
      y = store.spend
    )
  )
```
:::
::: {.column width="45%"}
\
\
\
\
```{r}
#| label: fig-distance-untransformed
#| fig-cap: Scatterplot of store spending vs distance to store
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(
    aes(
      x = distance.to.store,
      y = store.spend
    )
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
-   **Transforming variables**

```{r}
cor(customer$store.spend, 1 / sqrt(customer$distance.to.store))
```
\
```{r}
#| eval: false
#| code-line-numbers: "|2-4|8|"
customer |>
  mutate(
    distance.to.store_trans = 1 / sqrt(distance.to.store)
  ) |>
  ggplot() +
  geom_point(
    aes(
      x = distance.to.store_trans,
      y = store.spend
    )
  )
```
:::
::: {.column width="40%"}
\
\
\
\
```{r}
#| label: fig-distance-transformed
#| fig-cap: Scatterplot of store spending vs inverse square root distance
#| echo: false
#| lightbox: true
customer |>
  mutate(
    distance.to.store_trans = 1 / sqrt(distance.to.store)
  ) |>
  ggplot() +
  geom_point(
    aes(
      x = distance.to.store_trans,
      y = store.spend
    )
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Transforming variables**

    -   Understanding the logic behind inverse square root distance
```{r}
#| eval: false
#| code-line-numbers: "|2-6|"
ggplot() +
  geom_function(
    fun = function(x) {
      1 / sqrt(x)
    }
  ) +
  labs(
    x = "Distance",
    y = "Inverse square root distance"
  )
```
:::
::: {.column width="45%"}
\
\
\
```{r}
#| label: fig-inverse-sqrt-function
#| fig-cap: Using `geom_function()` to visualize inverse square root distance
#| echo: false
#| lightbox: true
ggplot() +
  geom_function(fun = function(x) {
    1 / sqrt(x)
  }) +
  labs(
    x = "Distance",
    y = "Inverse square root distance"
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualizing categorical variables**

    -   **Scatterplots: the tidyverse way**
```{r}
#| eval: false
#| code-line-numbers: "|3-8|"
customer |>
  ggplot() +
  geom_point(
    aes(
      x = sat.service,
      y = sat.selection
    )
  ) +
  labs(
    x = "Customer Satisfaction with Service",
    y = "Customer Satisfaction with Selection",
    title = "Customer as of June 2014"
  )
```
:::
::: {.column width="45%"}
\
\
\
```{r}
#| label: fig-categorical-overplotting
#| fig-cap: Using `geom_point()` with categorical variables (overplotting)
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(
    aes(
      x = sat.service,
      y = sat.selection
    )
  ) +
  labs(
    x = "Customer Satisfaction with Service",
    y = "Customer Satisfaction with Selection",
    title = "Customer as of June 2014"
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualizing categorical variables**

    -   **Scatterplots: the tidyverse way**
```{r}
#| eval: false
#| code-line-numbers: "|5-8|"
customer |>
  ggplot() +
  geom_point(
    aes(x = sat.service, y = sat.selection),
    position = position_jitter(
      width = 0.2,
      height = 0.2
    )
  ) +
  labs(
    x = "Customer Satisfaction with Service",
    y = "Customer Satisfaction with Selection",
    title = "Customer as of June 2014"
  )
```
:::
::: {.column width="45%"}
\
\
\
```{r}
#| label: fig-categorical-jitter
#| fig-cap: Using `position_jitter()` to reduce overplotting
#| echo: false
#| lightbox: true
customer |>
  ggplot() +
  geom_point(
    aes(x = sat.service, y = sat.selection),
    position = position_jitter(width = 0.2, height = 0.2)
  ) +
  labs(
    x = "Customer Satisfaction with Service",
    y = "Customer Satisfaction with Selection",
    title = "Customer as of June 2014"
  )
```
:::
::::

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/products/open-source/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**skimr**](https://CRAN.R-project.org/package=skimr), [**corrr**](https://CRAN.R-project.org/package=corrr) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}