---
title: "Comparing Groups: Tables and Visualizations"
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: ../000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: true
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: ../000_references/marketing_research_and_analytics.bib
csl: ../000_references/apa-7.csl
editor: source
---

```{r}
#| label: libraries
#| echo: false
library(tidyverse)
```

```{r}
#| label: palette
#| echo: false
umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   This presentation is based on [@chapman_r_2019, Chapter 5]

## 

- **Purpose**

    - Use descriptive summaries by groups and visualize them to investigate differences between groups

# Consumer segmentation survey

##

-   **age**: age of the consumer in years
-   **gender**: if the consumer is male of female
-   **income**: yearly disposable income of the consumer
-   **kids**: number of children of the consumer
-   **ownHome**: if the consumer owns a home
-   **subscribe**: if the consumer is subscribed or not
-   **Segment**: market segment assigned by a clustering algorithm [@chapman_r_2019, chapter 11], expert assignment or a segmentation typing tool

##

-   **Segment**:

    -   **Moving up**: consumers experiencing upward mobility in terms of their socioeconomic status
    -   **Suburb mix**: consumers living in suburban areas
    -   **Travelers**: consumers who prioritize experiences and adventures
    -   **Urban Hip**: consumers interested in urban culture, artistic expression, and modern trends

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|1-3|"
segmentation <- read_csv(
  file = "http://goo.gl/qw303p"
)

segmentation |>
  head(n = 5)
```
:::
::: {.column width="40%"}
-   **Import data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|2-8|"
segmentation <- segmentation |>
  mutate(
    gender = factor(gender, ordered = FALSE),
    kids = as.integer(kids),
    ownHome = factor(ownHome, ordered = FALSE),
    subscribe = factor(subscribe, ordered = FALSE),
    Segment = factor(Segment, ordered = FALSE)
  )

segmentation |>
  head(n = 5)
```
:::
::: {.column width="40%"}
-   **Transform data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="40%"}
```{r}
#| eval: false
?`+` # Arithmetic Operators

?formula # operators in a formula
```
:::
::: {.column width="60%"}
-   **Basic Formula Syntax**

    -   $\sim$ and $+$: operators

    -   $y$: response variable

    -   $x, z$: explanatory variables

    -   $y \sim x + z$: a formula which means that $y$ depends on $x$ and $z$

        -   $+$ is used to indicate the addition of predictor variables to the right of the formula
        -   Be careful not to confuse the arithmetic operator $+$ with $+$ within a formula
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|2-3|"
segmentation |>
  group_by(Segment, ownHome) |>
  summarise(mean_income = mean(income))
```
:::
::: {.column width="50%"}
-   **Descriptives for n-Way Groups**

    - the tidyverse way

      - Split data into $n$ subsets
      - Compute summary statistics
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|2-3|"
segmentation |>
  group_by(Segment, ownHome) |>
  summarise(sum_kids = sum(kids))
```
:::
::: {.column width="50%"}
-   **Descriptives for n-Way Groups**

    - the tidyverse way
    
      - Split data into $n$ subsets
      - Compute summary statistics
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
?`|` # Logical Operators

?lattice::xyplot # operators in a formula
# -------------- # you need first to install
# -------------- # the package lattice
```
:::
::: {.column width="50%"}
-   **Basic Formula Syntax**

    -   $\sim$, $+$ and $|$: operators

    -   $y$: response variable

    -   $x$: explanatory variable

    -   $z$: grouping variable

    -   $y \sim x | z$: $y$ depends on $x$ based on different groups defined by $z$

        -   $|$ is used to separate the grouping variable from the explanatory variable
        -   Be careful not to confuse the logical operator $|$ with $|$ within a formula
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|3-7|"
# Prepare data
subscriber_by_segment_home_ownership <- segmentation |>
  count(subscribe, Segment, ownHome) |>
  group_by(Segment, ownHome) |>
  mutate(n_pct = (n / sum(n)) * 100) |>
  ungroup() |>
  arrange(Segment, ownHome)

subscriber_by_segment_home_ownership
```
:::
::: {.column width="40%"}
-   **Visualization by group as frequencies**

    - the tidyverse way
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualization by group as frequencies**

    -   the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|4-8|"
subscriber_by_segment_home_ownership |>
  ggplot() +
  geom_col(aes(x = subscribe, y = n_pct)) +
  facet_wrap(
    facets = vars(Segment, ownHome),
    nrow = 2,
    ncol = 4
  )
```
:::
::: {.column width="45%"}
\
\
\
```{r}
#| label: fig-subscriber-by-segment-home-ownership
#| fig-cap: Visualization by group as frequencies using `facet_wrap()`
#| echo: false
#| lightbox: true
subscriber_by_segment_home_ownership |>
  ggplot() +
  geom_col(aes(x = subscribe, y = n_pct)) +
  facet_wrap(
    facets = vars(Segment, ownHome),
    nrow = 2,
    ncol = 4
  )
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualization by group as proportions**

    -   the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|3-6|"
# Prepare data
prop_table <- segmentation |>
  count(subscribe, Segment) |>
  group_by(Segment) |>
  mutate(n_pct = n / sum(n)) |>
  filter(subscribe == "subYes")
```
\
```{r}
#| eval: false
#| code-line-numbers: "|3-6|"
prop_table |>
  ggplot() +
  geom_col(
    aes(x = n_pct, y = Segment),
    fill = "darkolivegreen"
  ) +
  labs(
    x = "Subscriber proportion by Segment",
    y = NULL
  )
```
:::
::: {.column width="45%"}
\
\
\
\
\
\
\
\
```{r}
#| label: fig-subscriber-prop-by-segment
#| fig-cap: Visualization of subscriber proportions by segment
#| echo: false
#| lightbox: true
prop_table <- segmentation |>
  count(subscribe, Segment) |>
  group_by(Segment) |>
  mutate(n_pct = n / sum(n)) |>
  filter(subscribe == "subYes")

prop_table |>
  ggplot() +
  geom_col(
    aes(x = n_pct, y = Segment),
    fill = "darkolivegreen"
  ) +
  labs(x = "Subscriber proportion by Segment", y = NULL)
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualization by group with continuous data**

    -   the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|3-5|"
# Prepare data
seg_income_agg <- segmentation |>
  group_by(Segment, ownHome) |>
  summarise(mean_income = mean(income)) |>
  ungroup()
```
\
```{r}
#| eval: false
#| code-line-numbers: "|3-7|8|"
seg_income_agg |>
  ggplot() +
  geom_col(
    aes(x = Segment, y = mean_income, fill = ownHome),
    position = position_dodge(),
    color = "black"
  ) +
  scale_fill_manual(values = terrain.colors(n = 2))
```
:::
::: {.column width="45%"}
\
\
\
\
\
\
\
```{r}
#| label: fig-mean-income-by-segment-home-ownership
#| fig-cap: Mean income by segment and home ownership
#| echo: false
#| lightbox: true
seg_income_agg <- segmentation |>
  group_by(Segment, ownHome) |>
  summarise(mean_income = mean(income)) |>
  ungroup()

seg_income_agg |>
  ggplot() +
  geom_col(
    aes(x = Segment, y = mean_income, fill = ownHome),
    position = position_dodge(),
    color = "black"
  ) +
  scale_fill_manual(values = terrain.colors(n = 2))
```
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="55%"}
-   **Visualization by group with continuous data**

    -   the tidyverse way
```{r}
#| eval: false
#| code-line-numbers: "|3-4|"
segmentation |>
  ggplot() +
  geom_boxplot(aes(x = income, y = Segment)) +
  facet_wrap(facets = vars(ownHome)) +
  labs(x = "Income", y = NULL)
```
:::
::: {.column width="45%"}
\
\
\
```{r}
#| label: fig-income-by-segment-boxplot
#| fig-cap: Income distribution by segment faceted by home ownership
#| echo: false
#| lightbox: true
segmentation |>
  ggplot() +
  geom_boxplot(aes(x = income, y = Segment)) +
  facet_wrap(facets = vars(ownHome)) +
  labs(x = "Income", y = NULL)
```
:::
::::

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/products/open-source/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**lattice**](https://CRAN.R-project.org/package=lattice) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}