---
title: "Comparing Groups: Statistical Tests"
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: ../000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: true
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: ../000_references/marketing_research_and_analytics.bib
csl: ../000_references/apa-7.csl
editor: source
---

```{r}
#| label: libraries
#| echo: false
library(tidyverse)
library(skimr)
library(latex2exp)
library(tidymodels)
library(patchwork)
library(hardhat)
```

```{r}
#| label: palette
#| echo: false
umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   This presentation is based on [@chapman_r_2019, Chapter 6]

## 

- **Purpose**

    - Understand the use of statistical tests to identify differences between groups in data

# Consumer segmentation survey

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|1-3|"
segmentation <- read_csv(
  file = "http://goo.gl/qw303p"
)

segmentation |>
  head(n = 5)
```
:::
::: {.column width="40%"}
-   **Import data**
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "|2|"
segmentation |>
  count(Segment)
```
\
```{r}
#| code-line-numbers: "|3-7|"
segmentation |>
  count(subscribe, ownHome) |>
  pivot_wider(
    id_cols = subscribe,
    names_from = ownHome,
    values_from = n
  )
```
:::
::: {.column width="50%"}
-   Chi-squared test
:::
::::

## {.smaller .center}

-   Chi-squared test for given probabilities

$H_0: p_1 = \frac{1}{4} \land p_2 = \frac{1}{4} \land p_3 = \frac{1}{4} \land p_4 = \frac{1}{4}$

$H_1: p_1 \neq \frac{1}{4} \lor p_2 \neq \frac{1}{4} \lor p_3 = \frac{1}{4} \lor p_4 \neq \frac{1}{4}$

\

::: {.fragment}
$$
\begin{align}
  \chi^2 & = \sum_{i=1}^n \frac{(Observed_i - Expected_i)^2}{Expected_i} \\ 
  & = \frac{(70 - 300\frac{1}{4})^2}{300\frac{1}{4}} + \frac{(100 - 300\frac{1}{4})^2}{300\frac{1}{4}} +
                              \frac{(80 - 300\frac{1}{4})^2}{300\frac{1}{4}} + 
                              \frac{(50 - 300\frac{1}{4})^2}{300\frac{1}{4}}
  \end{align}
$$
:::

## {.smaller .center}

:::: {.columns}
::: {.column width="70%"}
-   Chi-squared test for given probabilities
```{r}
chi_statistic <- table(segmentation$Segment) |>
  chisq.test(
    p = c(1 / 4, 1 / 4, 1 / 4, 1 / 4)
  )

chi_statistic
```
\
```{r}
library(tidymodels)
segmentation |>
  chisq_test(
    response = Segment,
    p = c(1 / 4, 1 / 4, 1 / 4, 1 / 4)
  )
```
:::
::: {.column width="30%"}
\

-   **Base R way**
\
\
\
\
\
\
\
\
-   **tidymodels way**
:::
::::

## {.center}

```{r}
#| label: fig-chi-2-test-given-propabilities
#| fig-cap: Chi-squared test for given probabilities 
#| echo: false
#| lightbox: true
ggplot() +
  geom_function(
    fun = dchisq,
    args = list(df = 3, ncp = 0, log = FALSE),
    xlim = c(0, 20),
    color = '#2C3E50'
  ) +
  geom_ribbon(
    data = tibble(
      x = seq.int(
        from = qchisq(p = 0.05, df = 3, lower.tail = FALSE),
        to = 20,
        by = 0.01
      ),
      y = dchisq(x = x, df = 3)
    ),
    aes(x = x, ymin = 0, ymax = y),
    fill = '#E31A1C',
    alpha = 0.1
  ) +
  geom_vline(
    xintercept = qchisq(p = 0.05, df = 3, lower.tail = FALSE),
    color = "#E31A1C"
  ) +
  geom_vline(xintercept = unname(chi_statistic$statistic), color = "#18BC9C") +
  scale_x_continuous(
    breaks = c(
      0,
      qchisq(p = 0.05, df = 3, lower.tail = FALSE),
      unname(chi_statistic$statistic),
      20
    ),
    labels = scales::label_number(accuracy = 0.01)
  ) +
  labs(
    x = NULL,
    y = NULL,
    color = NULL,
    title = 'Chi-squared distribution function',
    subtitle = str_glue(
      'k=3
      Critical value: {qchisq(p = 0.05, df = 3, lower.tail = FALSE) |> round(digits=2)}
      Chi-squared statistic: {chi_statistic$statistic |> round(digits=2)}'
    )
  ) +
  theme(
    panel.border = element_rect(fill = NA, color = "black"),
    plot.background = element_rect(fill = "#f3fcfc"),
    panel.background = element_rect(fill = "#f3f7fc"),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = 'bottom',
    axis.text = element_text(face = "bold")
  )
```

## {.smaller .center}

-   Pearson's Chi-squared test

$H_0: p_{11} = \frac{260}{300}\frac{159}{300} \land p_{12} = \frac{260}{300}\frac{141}{300} \land p_{21} = \frac{40}{300}\frac{159}{300} \land p_{22} = \frac{40}{300}\frac{141}{300}$

$H_1: p_{11} \neq \frac{260}{300}\frac{159}{300} \lor p_{12} \neq \frac{260}{300}\frac{141}{300} \lor p_{21} \neq \frac{40}{300}\frac{159}{300} \lor p_{22} \neq \frac{40}{300}\frac{141}{300}$

\

::: {.fragment}
$$
\begin{align}
  \chi^2 & = \sum_{i=1}^n \frac{(Observed_i - Expected_i)^2}{Expected_i} \\
  & = \frac{(137 - 300\frac{260}{300}\frac{159}{300})^2}{300\frac{260}{300}\frac{159}{300}} + \frac{(123 - 300\frac{260}{300}\frac{141}{300})^2}{300\frac{260}{300}\frac{141}{300}} \\
  & \: \: \: \: \: + \frac{(22 - 300\frac{40}{300}\frac{159}{300})^2}{300\frac{40}{300}\frac{159}{300}} + \frac{(18 - 300\frac{40}{300}\frac{141}{300})^2}{300\frac{40}{300}\frac{141}{300}}
 \end{align}
$$
:::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
-   Pearson's Chi-squared test
```{r}
chi_statistic <- chisq.test(
  table(
    segmentation$subscribe,
    segmentation$ownHome
  ),
  correct = FALSE
)

chi_statistic
```
\
```{r}
segmentation |>
  chisq_test(
    formula = subscribe ~ ownHome,
    correct = FALSE
  )
```
:::
::: {.column width="40%"}
\
\

-   **Base R way**
\
\
\
\
\
\
\
\
\
-   **tidymodels way**
:::
::::

## {.center}

```{r}
#| label: fig-chi-2-test-pearson
#| fig-cap: Pearson's Chi-squared test 
#| echo: false
#| lightbox: true
ggplot() +
  geom_function(
    fun = dchisq,
    args = list(df = 1, ncp = 0, log = FALSE),
    xlim = c(0, 20),
    color = '#2C3E50'
  ) +
  geom_ribbon(
    data = tibble(
      x = seq.int(
        from = qchisq(p = 0.05, df = 1, lower.tail = FALSE),
        to = 20,
        by = 0.01
      ),
      y = dchisq(x = x, df = 1)
    ),
    aes(x = x, ymin = 0, ymax = y),
    fill = '#E31A1C',
    alpha = 0.1
  ) +
  geom_vline(
    xintercept = qchisq(p = 0.05, df = 1, lower.tail = FALSE),
    color = "#E31A1C"
  ) +
  geom_vline(xintercept = unname(chi_statistic$statistic), color = "#18BC9C") +
  scale_x_continuous(
    breaks = c(
      qchisq(p = 0.05, df = 1, lower.tail = FALSE),
      unname(chi_statistic$statistic),
      20
    ),
    labels = scales::label_number(accuracy = 0.01)
  ) +
  labs(
    x = NULL,
    y = NULL,
    color = NULL,
    title = 'Chi-squared distribution function',
    subtitle = str_glue(
      'k=1
      Critical value: {qchisq(p = 0.05, df = 1, lower.tail = FALSE) |> round(digits=2)}
      Chi-squared statistic: {chi_statistic$statistic |> round(digits=2)}'
    )
  ) +
  theme(
    panel.border = element_rect(fill = NA, color = "black"),
    plot.background = element_rect(fill = "#f3fcfc"),
    panel.background = element_rect(fill = "#f3f7fc"),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = 'bottom',
    axis.text = element_text(face = "bold")
  )
```

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   2 sample t-test: independent samples

```{r}
#| eval: false
segmentation |>
  ggplot() +
  geom_histogram(
    aes(x = income),
    color = 'black'
  ) +
  facet_wrap(facets = vars(ownHome))
```
\
\
\
\
```{r}
#| eval: false
segmentation |>
  group_by(ownHome) |>
  summarise(
    mean_income = mean(income),
    var_income = var(income),
    n = n()
  )
```
:::
::: {.column width="50%"}
\
```{r}
#| label: fig-2-sample-test-comparing-means
#| fig-cap: Visualizing distributions to compare means in 2 samples
#| echo: false
#| lightbox: true
segmentation |>
  ggplot() +
  geom_histogram(
    aes(x = income),
    color = 'black'
  ) +
  facet_wrap(facets = vars(ownHome))
```
```{r}
#| echo: false
segmentation |>
  group_by(ownHome) |>
  summarise(
    mean_income = mean(income),
    var_income = var(income),
    n = n()
  )
```
:::
::::

## {.center .smaller}

-   2 sample t-test: independent samples

$H_0: \mu_{ownNo} - \mu_{ownYes}= 0$ 

$H_1: \mu_{ownNo} - \mu_{ownYes} \neq 0$

\

::: {.fragment}
$$
\begin{aligned}
t  & = \frac{\overline{ownNo} - \overline{ownYes}}{\sqrt{\frac{s_{ownNo}^2}{n_{ownNo}} - \frac{s_{ownYes}^2}{n_{ownYes}}}} \\
& = \frac{47391.01 - 54934.68}{\sqrt{ \frac{358692875}{159} - \frac{430890091}{141}}} \\ 
& \approx -3.273094
\end{aligned}
$$
:::

## {.smaller .center}

::::: {.columns}

:::: {.column style="width: 70%; font-size: 85%;"}
-   2 sample t-test: independent samples
```{r}
#| code-line-numbers: "|3|"
t.test(
  formula = income ~ ownHome,
  data = segmentation,
  alternative = 'two.sided',
  mu = 0,
  conf.level = 0.95
)
```
```{r}
#| code-line-numbers: "|1|5|"
#| classes: fragment
segmentation |>
  t_test(
    formula = income ~ ownHome,
    alternative = "two-sided",
    order = c("ownNo", "ownYes"),
    mu = 0,
    conf_level = 0.95
  )
```
::::

:::: {.column width="30%"}
\

-   **R base way**
\
\
\
\
\
\
\
\
\
\

::: {.fragment}
-   **tidymodels way**
:::

::::

:::::

## {.center}

```{r}
#| label: fig-2-sample-t-test-independent-samples
#| fig-cap: "2 sample t-test: independent samples" 
#| echo: false
#| lightbox: true
t_test <- t.test(
  formula = income ~ ownHome,
  data = segmentation,
  alternative = 'two.sided',
  mu = 0,
  conf.level = 0.95
)

ggplot() +
  geom_function(
    fun = dt,
    args = list(df = t_test$parameter),
    xlim = c(-5, 5),
    color = '#2C3E50'
  ) +
  geom_vline(
    xintercept = qt(
      p = 0.025,
      df = t_test$parameter,
      ncp = 0,
      lower.tail = TRUE
    ),
    color = "#E31A1C"
  ) +
  geom_vline(
    xintercept = qt(
      p = 0.025,
      df = t_test$parameter,
      ncp = 0,
      lower.tail = FALSE
    ),
    color = "#E31A1C"
  ) +
  geom_vline(xintercept = unname(t_test$statistic), color = "#18BC9C") +
  geom_ribbon(
    data = tibble(
      x = seq.int(
        from = -5,
        to = qt(p = 0.025, df = t_test$parameter, ncp = 0, lower.tail = TRUE),
        by = 0.01
      ),
      y = dt(x = x, df = t_test$parameter, ncp = 0)
    ),
    aes(x = x, ymin = 0, ymax = y),
    fill = '#CCBE93',
    alpha = 0.5
  ) +
  geom_ribbon(
    data = tibble(
      x = seq.int(
        from = qt(
          p = 0.025,
          df = t_test$parameter,
          ncp = 0,
          lower.tail = FALSE
        ),
        to = 5,
        by = 0.01
      ),
      y = dt(x = x, df = t_test$parameter, ncp = 0)
    ),
    aes(x = x, ymin = 0, ymax = y),
    fill = '#CCBE93',
    alpha = 0.5
  ) +
  scale_x_continuous(
    breaks = c(
      -5,
      unname(t_test$statistic),
      qt(p = 0.025, df = t_test$parameter, ncp = 0, lower.tail = TRUE),
      qt(p = 0.025, df = t_test$parameter, ncp = 0, lower.tail = FALSE),
      5
    ),
    labels = scales::label_number(accuracy = 0.01)
  ) +
  labs(
    x = NULL,
    y = NULL,
    color = NULL,
    title = "Student's t-distribution distribution function",
    subtitle = str_glue(
      'Degrees of freedom= {t_test$parameter |> round(digits=2)},
      Critical values: ({qt(p = 0.025, df = t_test$parameter, ncp = 0,
                         lower.tail = TRUE) |> round(digits=2)},{qt(p = 0.025, df = t_test$parameter, ncp = 0,
                         lower.tail = FALSE) |> round(digits=2)})
                         T statistic: {unname(t_test$statistic) |> round(digits=2)}'
    )
  ) +
  theme(
    panel.border = element_rect(fill = NA, color = "black"),
    plot.background = element_rect(fill = "#f3fcfc"),
    panel.background = element_rect(fill = "#f3f7fc"),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = 'bottom',
    axis.text = element_text(face = "bold")
  )
```

## {.smaller .center}

-   2 sample t-test: independent samples

    -   Confidence interval

        $$
        c_L < \mu_{ownNo} - \mu_{ownYes} < c_U
        $$

        -   $\mu_{ownNo} - \mu_{ownYes}$ is not a random variable, so we need to use a random variable

        $$
        P \left( t_L < \frac{(\overline{x}_{ownNo} - \overline{x}_{ownYes}) - (\mu_{ownNo} - \mu_{ownYes})}{\sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } +\frac{s^2_{ownYes}}{n_{ownYes}}}} < t_U \right) = 0.95
        $$

        -   $\overline{x}_{ownNo} - \overline{x}_{ownYes}$ is a random variable

## {.smaller .center}

-   2 sample t-test: independent samples

    -   Confidence interval

        -   $\frac{(\overline{x}_{ownNo} - \overline{x}_{ownYes}) - (\mu_{ownNo} - \mu_{ownYes})}{\sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } +\frac{s^2_{ownYes}}{n_{ownYes}}}}$ is also a random variable with student's t-distribution and $\nu \approx \frac{\left(\frac{s_{ownNo}^2}{n_{ownNo}} + \frac{s_2^2}{n_{ownYes}}\right)^2}{\frac{\left(\frac{s_{ownNo}^2}{n_{ownNo}}\right)^2}{n_{ownNo}-1} + \frac{\left(\frac{s_2^2}{n_{ownYes}}\right)^2}{n_{ownYes}-1}} = 285.2521$ degrees of freedom

        -   Also we need to specify $t_L$ and $t_U$

            ```{r}
            t_L <- qt(p = 0.025, df = 285.25, lower.tail = TRUE)
            t_L
            t_U <- qt(p = 0.975, df = 285.25, lower.tail = TRUE)
            t_U
            ```

## {.smaller .center}

-   2 sample t-test: independent samples

    -   Confidence interval

        ::: {style="font-size: 75%;"}

        $$
        \begin{alignedat}{2}
        & P \left( (\overline{x}_{ownNo} - \overline{x}_{ownYes}) - t_U \times \sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } + \frac{s^2_{ownYes}}{n_{ownYes}}} \right. \\
        & \left. \quad \quad < \mu_{ownNo} - \mu_{ownYes} < (\overline{x}_{ownNo} - \overline{x}_{ownYes}) - t_L \times \sqrt{\frac{s^2_{ownNo}}{n_{ownNo} } + \frac{s^2_{ownYes}}{n_{ownYes}}} \right) &&= 0.95 \\[1.5ex]
        & P \left( -7543.674 - 1.968315 \times 2304.753 \right. \\
        & \left. \quad \quad < \mu_{ownNo} - \mu_{ownYes} < -7543.674 + 1.968315 \times 2304.753 \right) &&= 0.95 \\[1.5ex]
        & P \left( -12080.16 < \mu_{ownNo} - \mu_{ownYes} < -3007.193 \right) &&= 0.95
        \end{alignedat}
        $$

        :::

        -   In the long run 95% of confidence intervals constructed in this manner will contain the true parameter

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   Testing Multiple Group Means

    - Analysis of Variance (ANOVA)
```{r}
#| code-line-numbers: "|3-5|"
#| classes: fragment
segmentation |>
  summarise(
    mean = mean(income),
    variance = var(income),
    n = n()
  )
```
```{r}
#| code-line-numbers: "|2|4-6|"
#| classes: fragment
segmentation |>
  group_by(Segment) |>
  summarise(
    mean = mean(income),
    variance = var(income),
    n = n()
  )
```
:::
::: {.column width="50%"}
\
\
\
\
\
```{r}
#| label: fig-visualization-anova
#| fig-cap: "Visualizing Analysis of Variance (ANOVA)" 
#| echo: false
#| lightbox: true
segmentation |>
  ggplot(aes(x = Segment, y = income)) +
  geom_point(position = position_jitter(width = 0.1)) +
  geom_hline(yintercept = mean(segmentation$income), color = '#2C3E50') +
  stat_summary(
    geom = "point",
    fun = 'mean',
    size = 4,
    shape = 21,
    fill = '#E31A1C',
    color = 'black'
  )
```
:::
::::

## {.smaller}

-   Testing Multiple Group Means: Analysis of Variance (ANOVA)

$H_0: \mu_{Moving\;up} = \mu_{Suburb\;mix} = \mu_{Travelers} = \mu_{Urban\;hip}$

$H_1: \text{At least one group mean is different from the rest}$

\

::: {.fragment}
$n = \sum_{j=1}^4 n_j = n_1 + \cdots + n_4 = 70 + 100 + 80 + 50 = 300$

$\overline{income} = \frac{1}{n} \sum_{j=1}^4 \sum_{i=1}^{n_j} income_{ij} = 50936.54$

$\overline{income}_j = \frac{1}{n_j} \sum_{i=1}^{n_j} income_{ij}$
:::

\

::: {.fragment}
$$
F = \frac{\frac{\sum_{j=1}^4 \sum_{i=1}^{n_j} (\overline{income}_j - \overline{income})^2}{4-1}}{\frac{\sum_{j=1}^4 \sum_{i=1}^{n_j} (income_{ij} - \overline{income}_j)^2}{300 - 4}} = \frac{\frac{54969675428}{3}}{\frac{66281072794}{296}} = \frac{18323225143}{223922543} = 81.82841
$$
:::

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
```{r}
#| code-line-numbers: "|1-4|5|6|"
anova_table <- aov(
  data = segmentation,
  formula = income ~ Segment
) |>
  anova() |>
  tidy()

anova_table
```
:::
::: {.column width="40%"}
-   Testing Multiple Group Means

    -  Analysis of Variance (ANOVA)

       -   **tidymodels way**
:::
::::

## {.center}

```{r}
#| label: fig-anova
#| fig-cap: "Testing Multiple Group Means: Analysis of Variance (ANOVA)" 
#| echo: false
#| lightbox: true
anova_table <- aov(
  data = segmentation,
  formula = income ~ Segment
) |>
  anova()

n <- nrow(segmentation)
k <- segmentation |>
  count(Segment) |>
  nrow()
df1 <- k - 1
df2 <- n - k

ggplot() +
  geom_function(
    fun = df,
    args = list(df1 = df1, df2 = df2, log = FALSE),
    xlim = c(0, 90),
    color = '#2C3E50'
  ) +
  geom_ribbon(
    data = tibble(
      x = seq.int(
        from = qf(p = 0.05, df1 = df1, df2 = df2, lower.tail = FALSE),
        to = 90,
        by = 0.01
      ),
      y = df(x = x, df1 = df1, df2 = df2)
    ),
    aes(x = x, ymin = 0, ymax = y),
    fill = '#E31A1C',
    alpha = 0.1
  ) +
  geom_vline(
    xintercept = qf(p = 0.05, df1 = df1, df2 = df2, lower.tail = FALSE),
    color = "#E31A1C"
  ) +
  geom_vline(xintercept = unname(anova_table$`F value`)[1], color = "#18BC9C") +
  scale_x_continuous(
    breaks = c(
      qf(p = 0.05, df1 = df1, df2 = df2, lower.tail = FALSE),
      unname(anova_table$`F value`)[1],
      90
    ),
    labels = scales::label_number(accuracy = 0.1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    color = NULL,
    title = 'F-squared distribution function',
    subtitle = str_glue(
      'df1={df1},
      df2={df2},
      Critical value: {qf(p = 0.05, df1 = df1, df2=df2, lower.tail = FALSE) |> round(digits=2)}
      F statistic: {unname(anova_table$`F value`)[1] |> round(digits=2)}'
    )
  ) +
  theme(
    panel.border = element_rect(fill = NA, color = "black"),
    plot.background = element_rect(fill = "#f3fcfc"),
    panel.background = element_rect(fill = "#f3f7fc"),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = 'bottom',
    axis.text = element_text(face = "bold")
  )
```

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   Testing Multiple Group Means
    
    -  Analysis of Variance (ANOVA)
```{r}
segmentation |>
  distinct(Segment) |>
  arrange(Segment) |>
  rowid_to_column(var = 'i')
```
```{r}
segmentation |>
  distinct(ownHome) |>
  rowid_to_column(var = 'j')
```
:::
::: {.column width="50%"}
\
\

-  By `Segment`

\
\
\
\
\

-  By `ownHome` 
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   Testing Multiple Group Means
    
    -  Analysis of Variance (ANOVA)
```{r}
segmentation |>
  count(
    Segment,
    ownHome,
    name = "n_ij"
  )
```
:::
::: {.column width="50%"}
\
\

-  Number of elements in each group

   -  $n_{11} = 47, n_{12} = 23$ 
   -  $n_{21} = 52, n_{22} = 48$
   -  $n_{31} = 20, n_{32} = 60$
   -  $n_{41} = 40, n_{42} = 10$
:::
::::

## {.smaller .center}

:::: {.columns}
::: {.column width="50%"}
-   Testing Multiple Group Means
    
    -  Analysis of Variance (ANOVA)
```{r}
mu_ij <- segmentation |>
  group_by(Segment, ownHome) |>
  summarise(mean = mean(income)) |>
  ungroup()
mu_ij
```
```{r}
segmentation |>
  select(income, Segment, ownHome) |>
  head(n = 2)
```
:::
::: {.column style="width: 50%; font-size: 75%;"}
\
\
\
\

-  Means

   -  $\mu_{11} = 54497.68, \mu_{12} = 50216.37$ 
   -  $\mu_{21} = 54932.83, \mu_{22} = 55143.21$
   -  $\mu_{31} = 63188.42, \mu_{32} = 61889.12$
   -  $\mu_{41} = 21337.59, \mu_{42} = 23059.27$

\
\
\
\

-  Incomes

   - $income_{211} = 49482.81$
   - $income_{221} = 35546.29$
   - $\phantom{income_{ijk}}\:\:\: \vdots$
:::
::::

## {.center}

```{r}
#| label: fig-visualizing-anova-2-categories
#| fig-cap: "Visualizing Analysis of Variance (ANOVA) for 2 variables" 
#| echo: false
#| lightbox: true

#| echo: false
mean_Segment <- segmentation |>
  group_by(Segment) |>
  summarise(mean = mean(income)) |>
  ungroup()

mean_ownHome <- segmentation |>
  group_by(ownHome) |>
  summarise(mean = mean(income)) |>
  ungroup()

mean_Segment_ownHome <- segmentation |>
  group_by(Segment, ownHome) |>
  summarise(mean = mean(income)) |>
  ungroup()

g1 <- segmentation |>
  ggplot(aes(x = Segment, y = income)) +
  geom_point(position = position_jitter(width = 0.1)) +
  stat_summary(
    geom = "point",
    fun = 'mean',
    size = 3,
    shape = 21,
    fill = '#E31A1C',
    color = 'black'
  ) +
  geom_line(
    data = mean_Segment,
    aes(x = Segment, y = mean, group = 1),
    color = "#2C3E50"
  )

g2 <- segmentation |>
  ggplot(aes(x = ownHome, y = income)) +
  geom_point(position = position_jitter(width = 0.1)) +
  stat_summary(
    geom = "point",
    fun = 'mean',
    size = 3,
    shape = 21,
    fill = '#E31A1C',
    color = 'black'
  ) +
  geom_line(
    data = mean_ownHome,
    aes(x = ownHome, y = mean, group = 1),
    color = "#2C3E50"
  )

g3 <- segmentation |>
  ggplot(aes(x = Segment, y = income, fill = ownHome, group = ownHome)) +
  geom_point(position = position_jitter(width = 0.1), shape = 21) +
  geom_line(
    data = mean_Segment_ownHome,
    aes(x = Segment, y = mean, color = ownHome)
  ) +
  stat_summary(
    geom = "point",
    fun = 'mean',
    size = 3,
    shape = 21,
    fill = '#E31A1C',
    color = 'black'
  ) +
  scale_color_manual(values = c("#18BC9C", "#CCBE93")) +
  scale_fill_manual(values = c("#18BC9C", "#CCBE93")) +
  theme(legend.position = "bottom")

(g1 / g2) | g3
```

## {.smaller .center}

-   Testing Multiple Group Means: Analysis of Variance (ANOVA)

$$
\begin{split}
  income_{ijk} =  & \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk} \\ 
  & \text{ where } + \epsilon_i \sim \mathcal{N}(0, \sigma^2) \\ 
  & \text{ and } i = 1, 2, 3, 4 \\
  & j = 1, 2 \\
  & k = 1, \ldots n_{ij} \\
  & \mu = \mu_{11} \\
  & \alpha_1 = \beta_1 = 0 \\
  & (\alpha\beta)_{11} = (\alpha\beta)_{12} = 0 \\ 
  & (\alpha\beta)_{21} = (\alpha\beta)_{31} = (\alpha\beta)_{41} = 0 \\  
  \end{split}
$$

## {.smaller .center}

-   Testing Multiple Group Means: Analysis of Variance (ANOVA)

$$
\begin{split}
  \widehat{income}_{ijk} & = \widehat{\mu} + \widehat{\alpha}_i + \widehat{\beta}_j + (\widehat{\alpha\beta})_{ij} + \widehat{\epsilon}_{ijk} \\ 
  & \text{ and } i = 1, 2, 3, 4 \\
  & j = 1, 2 \\
  & k = 1, \ldots n_{ij} \\
  & \widehat{\mu} = \widehat{\mu}_{11} \\
  & \widehat{\alpha}_2, \widehat{\alpha}_3, \widehat{\alpha}_4 \\
  & \widehat{\beta}_2 \\ 
  & (\widehat{\alpha\beta})_{22}, (\widehat{\alpha\beta})_{32}, (\widehat{\alpha\beta})_{42}
  \end{split}
$$

$$
income_{ijk} - \widehat{income}_{ijk} = \widehat{\epsilon}_{ijk}
$$

## {.smaller .center}

::: columns
::: {.column width="56%"}
-   Testing Multiple Group Means

    -  Analysis of Variance (ANOVA)
```{r}
#| eval: false
segmentation |>
  select(
    income,
    Segment,
    ownHome
  ) |>
  head(n = 2)
```
\
```{r}
#| eval: false
framed <- model_frame(
  formula = income ~ Segment + ownHome + Segment:ownHome,
  data = segmentation
)

model_matrix(
  terms = framed$terms,
  data = framed$data
) |>
  head(n = 2) |>
  glimpse()
```
:::

::: {.column width="44%"}
\
\

-  Data
```{r}
#| echo: false
segmentation |>
  select(
    income,
    Segment,
    ownHome
  ) |>
  head(n = 2)
```
\

-  Model matrix
```{r}
#| echo: false
framed <- model_frame(
  formula = income ~ Segment + ownHome + Segment:ownHome,
  data = segmentation
)

model_matrix(
  terms = framed$terms,
  data = framed$data
) |>
  head(n = 2) |>
  glimpse()
```
:::
:::

## {.smaller}

-   Testing Multiple Group Means: Analysis of Variance (ANOVA)

    -   Model

        $$
        \begin{bmatrix}
          49482.81 \\
          35546.29 \\
          \vdots
          \end{bmatrix} =
          \begin{bmatrix}
          1 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
          1 & 1 & 0 & 0 & 1 & 1 & 0 & 0 \\
          \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots
          \end{bmatrix} 
          \begin{bmatrix}
          \mu \\
          \alpha_2 \\
          \alpha_3 \\
          \alpha_4 \\
          \beta_2 \\
          (\alpha\beta)_{22} \\
          (\alpha\beta)_{32} \\
          (\alpha\beta)_{42} \\
          \end{bmatrix}
        $$

    -   Coefficients to estimate using `aov`

        $$
        \widehat{\mu} = \widehat{\mu}_{11}, \widehat{\alpha}_2, \widehat{\alpha}_3, \widehat{\alpha}_4, \widehat{\beta}_2, (\widehat{\alpha\beta})_{22}, (\widehat{\alpha\beta})_{32}, (\widehat{\alpha\beta})_{42}
        $$

## {.smaller .center}

:::: {.columns}
::: {.column width="60%"}
-   Testing Multiple Group Means

    -  Analysis of Variance (ANOVA)
```{r}
#| echo: true
model_aov <- aov(
  formula = income ~ Segment + ownHome + Segment:ownHome,
  data = segmentation
)

coef(model_aov) |>
  enframe(name = "coef")
```
:::
::: {.column style="width: 40%; font-size: 75%;"}
\
\
\

-  Estimated coefficients

   -  $\widehat{\mu} = \widehat{\mu}_{11} = 54497.682$
   -  $\widehat{\alpha}_2 = 435.152$
   -  $\widehat{\alpha}_3 = 8690.738$
   -  $\widehat{\alpha}_4 = -33160.091$
   -  $\widehat{\beta}_2 = -4281.313$ 
   -  $(\widehat{\alpha\beta})_{22} = 4491.690$
   -  $(\widehat{\alpha\beta})_{32} = 2982.007$ 
   -  $(\widehat{\alpha\beta})_{42} = 6002.996$
:::
::::

## {.smaller}

-   Testing Multiple Group Means: Analysis of Variance (ANOVA)

    -   Segment

        $H_0: \mu_{Moving\;up} = \mu_{Suburb\;mix} = \mu_{Travelers} = \mu_{Urban\;hip}$

        $H_1: \text{At least one group mean is different from the rest}$

    -   ownHome

        $H_0: \mu_{ownNo} = \mu_{ownYes}$
        
        $H_1: \text{At least one group mean is different from the rest}$

    -   Segment, ownHome

        $\begin{align}
          H_0: & \mu_{Moving\;up,\;ownNo} - \mu_{Moving\;up,\;ownYes} = \\
               & \mu_{Suburb\;mix,\;ownNo} - \mu_{Suburb\;mix,\;ownYes} = \\ 
               & \mu_{Travelers,\;ownNo} - \mu_{Travelers,\;ownYes} = \\
               & \mu_{Urban\;hip,\;ownNo} - \mu_{Urban\;hip,\;ownYes}
              \end{align}$
        
        $H_1: \text{At least one difference group mean is different from the rest}$

## {.smaller .center}

:::: {.columns}
::: {.column width="30%"}
```{r}
#| code-line-numbers: '|2|'
#| eval: false
model_aov |>
  anova() |>
  tidy()
```
:::
::: {.column width="70%"}
-   Testing Multiple Group Means

    -  Analysis of Variance (ANOVA)
```{r}
#| echo: false
model_aov |>
  anova() |>
  tidy()
```
:::
::::

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/products/open-source/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**skimr**](https://CRAN.R-project.org/package=skimr), [**latex2exp**](https://CRAN.R-project.org/package=latex2exp), [**tidymodels**](https://CRAN.R-project.org/package=tidymodels), [**patchwork**](https://CRAN.R-project.org/package=patchwork), [**hardhat**](https://CRAN.R-project.org/package=hardhat) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}